---
title: "Citing a Set of Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citing a Set of Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This article picks up where [S3 Setup and Promotion](start-on-s3.html) left off.
You should have a study's data onboarded and readable before starting here.

Output below is copied from a real run against a sandbox bucket, with long paths
and wide tables shortened. Project names, hashes and bucket will not match yours.

## The problem

You built a result from a few tables. Months later, more data has arrived and
several of those tables have newer versions. Someone asks which data produced the
result.

That question has no answer unless somebody wrote the versions down. Usually
nobody did.

## What a set is

A **set** is a citable list of exact data versions. It has a name and a version
of its own, so one string refers to the whole collection and resolves to exactly
those tables at exactly those versions -- next month, next year.

It stores pointers, not data, so making one is cheap. And it does not drift: a set
pins versions, and never means "whatever is current".

## Where you left off

The previous article left you with a `store` for your bucket, a `conn` to the
repo that onboards data, and `dev_dir`, its working directory.

You are about to have two connections, so rename that one to say which it is:

```{r}
imported <- conn   # the connection to the onboarding repo
```

## A repository for the product

The onboarding repo takes files in and versions them. Your product is a different
job -- it *cites* that data -- so it gets its own repo, next to the first one in
the same bucket:

```
s3://study-001-datom/
    imported/datom/     the onboarded tables      (previous article)
    adam/datom/         the product               (this article)
```

`imported` is the word datom uses for a table that came from a file, as opposed
to `derived` for one written from R. `raw/`, `edc/` or `sdtm/` work just as well
-- the prefix is a plain string.

```{r}
product_dir <- fs::path(tempdir(), "study_001_adam")

product_data <- datom_store_s3(
  bucket     = "study-001-datom",
  prefix     = "adam/",
  region     = "us-east-1",
  access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
  secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY")
)

product_store <- datom_store(
  data       = product_data,
  github_pat = Sys.getenv("GITHUB_PAT")
)

datom_init_repo(
  path         = product_dir,
  project_name = "STUDY_ADAM",
  store        = product_store,
  create_repo  = TRUE,
  repo_name    = "study-001-adam",
  mode         = "product",          # this repo builds; it does not onboard files
  set          = "trial_product"     # the set it owns
)

product <- datom_get_conn(path = product_dir, store = product_store)
print(product)
#> -- datom connection
#> * Project: "STUDY_ADAM"
#> * Backend: "s3"
#> * Role: "developer"
#> * Data prefix: "adam/"
#> * Governance: not attached
#> * Data repo: <https://github.com/.../study-001-adam.git>
```

One repo owns one set, and it names that set here. You never pass the name again.

## Build the set

See what the onboarding repo holds:

```{r}
datom_list(imported)
#>   name  kind current_version current_data_sha         last_updated
#> 1   dm table        9a256607         e547f03d 2026-09-26T01:00:33Z
#> 2   ae table        ddddfe5a         d5f8dd5a 2026-09-26T01:00:28Z
#> 3   ex table        419b071c         ab96afc3 2026-09-26T01:00:37Z
#> 4   lb table        1319d915         5d419c60 2026-09-26T01:00:41Z
```

Pick the ones your product used -- a set cites what you used, not everything
available -- and note their versions:

```{r}
dm_v <- datom_history(imported, "dm")$version[1]
ex_v <- datom_history(imported, "ex")$version[1]
```

Then derive whatever your product produces, and write it into the **product's**
repo:

```{r}
dm   <- datom_read(imported, "dm")
adsl <- dm[, c("STUDYID", "USUBJID", "AGE", "SEX")]   # your real derivation here

datom_write(product, data = adsl, name = "adsl")
adsl_v <- datom_history(product, "adsl")$version[1]
```

Now write the set, citing the inputs you used **and** the table you produced:

```{r}
datom_write_set(
  product,
  members = list(
    datom_member(imported, "dm", dm_v,     tags = list(type = "input")),
    datom_member(imported, "ex", ex_v,     tags = list(type = "input")),
    datom_member(product,  "adsl", adsl_v, tags = list(type = "output"))
  ),
  tags = list(description = "Trial data cut, month 1")
)
```

`datom_member(imported, ...)` is how a set cites another project's table: you
resolve the version through that project's connection, and the pointer remembers
where it came from. `datom_member(product, ...)` does the same for a table in the
product's own repo.

**The set is the product's index, so put everything the product produced in it.**
A derived table sitting in the product's storage that the set does not cite is
orphaned -- nothing points at it, so nothing will find it later. If you wrote it,
cite it.

Labels like `type = "input"` are your own words, and they are about to earn their
keep: they are how you refresh your inputs without touching your outputs.

## Cite it, and read it somewhere else

The version that came back is the citation. Put `9e0c05e6` in a report and it
resolves to those two tables forever.

Reading it needs access to the product's storage and **nothing else** -- no git
clone, no GitHub token:

```{r}
reader_store <- datom_store(
  data = datom_store_s3(
    bucket     = "study-001-datom",
    prefix     = "adam/",
    region     = "us-east-1",
    access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
    secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY")
  )
  # no github_pat: that absence is what makes this a reader
)
reader <- datom_get_conn(store = reader_store, project_name = "STUDY_ADAM")

x <- datom_get_set(reader, "trial_product")
print(x)
```

That prints the set's name, project and version, then each member with its
labels -- the two inputs and the output you produced.

`datom_list_members(x)` gives the same thing as a data frame, one row per label,
if you would rather filter it than read it.

Reading the set gave you pointers. For the data, ask for one member through a
connection to **that member's** project:

```{r}
head(datom_fetch_member(imported, x, "dm"))
#> # A tibble: 6 x 12
#>   STUDYID   DOMAIN USUBJID         SUBJID   AGE AGEU  SEX   ...
#> 1 STUDY-001 DM     STUDY-001-001        1    71 YEARS F
#> 2 STUDY-001 DM     STUDY-001-002        2    27 YEARS M
```

That split is deliberate, and it is what makes a set worth having: the set is
citable by someone who has no access to the data inside it.

## An input moves

More data arrives and the onboarding repo takes it in, as before:

```{r}
datom_sync(imported, datom_sync_manifest(imported))
#> i Scanned 4 files: 0 new, 1 changed, 3 unchanged.
#> v Wrote "dm" (full): "dfd8e34c"
#> i Sync complete: 1 succeeded, 0 failed, 3 skipped.
```

Your set still points at the older version, which is correct -- a citation does
not move on its own. Move it when you decide to, and say which members you mean:

```{r}
x <- datom_get_set(product, "trial_product")
x <- datom_update_members(x, imported, tags = list(type = "input"))
#> project STUDY_001:
#>   dm  9a256607 -> dfd8e34c
#> i Nothing has been written. Write the set with `datom_write_set(conn, x)`.
```

It opens with a count of how many of the selected members moved, then names each
one, `old -> new`, grouped by project.

This is where the labels pay off. `tags = list(type = "input")` refreshes your
inputs and leaves your output alone, which is what you want -- your `adsl` moves
when you rebuild it, not when an input changes underneath it.

It also keeps this to one connection. Without the filter, every member is
selected, including `adsl` in the product's own project, and datom would ask for a
connection for that project too rather than quietly skipping it.

It reports what moved, `old -> new`, **before** anything is written. You are
holding an edited set; the write is yours to make. Labels came along with the
pointer.

```{r}
datom_write_set(product, x)
```

Run the same refresh when nothing has changed and it costs nothing:

```{r}
datom_write_set(product, datom_update_members(datom_get_set(product, "trial_product"),
                                              imported, tags = list(type = "input")))
#> i No changes detected for set "trial_product". Skipping write.
```

No new version. A set's version moves when its content moves, so re-running a
build script is free.

## Retire a member

```{r}
x <- datom_get_set(product, "trial_product")
x <- datom_remove_members(x, "ex")     # an input you no longer use
datom_write_set(product, x)
```

Select by name or by label rather than by position, because positions shift when
someone adds a member.

Edits made together are written together. Repoint and remove before writing, and
the commit message says so:

```{r}
datom_history(product, "trial_product")[, c("version", "commit_message")]
#>    version                                          commit_message
#> 1 31d711b2  Update trial_product: repoint 1 member, drop 1 member
#> 2 ae193a04  Update trial_product
```

## What a set does not do

* It holds no data. Removing a member deletes nothing -- the table it pointed at
  is untouched, and older versions of the set still cite it.
* It does not drift. A member pins one version, chosen by you.
* It does not change its members. Writing a set is read-only on them.

Two more things worth knowing before you build one in anger:

**A write replaces the set's current version**, because one repo owns one set.
Earlier versions stay readable by their own version string, so nothing is lost --
but `datom_get_set(conn, name)` always gives you the newest.

**Labels are content.** Changing a description or a member's label mints a new
version, the same as changing a member would. That is the point: the set now says
something different about itself.

## Where you are

You have a study's data in one repo and a product in another that cites exact
versions of it. The product's version is one string you can put in a report, and
it will resolve to the same tables in a year.

A product can also pool several studies from several buckets, which needs
credentials that reach all of them. That is a separate article.

## Teardown

If you were following along on real infrastructure:

```{r}
datom_repo_delete(product,  confirm = "STUDY_ADAM")
datom_repo_delete(imported, confirm = "STUDY_001")
```

The buckets are yours; datom removes only the namespaces it created.
