---
title: "Citing a Set of Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citing a Set of Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This article follows on from [S3 Setup and Promotion](start-on-s3.html). It assumes
you already have a study's data onboarded and can read it back. If you have not
done that, start there.

Output shown below is from a real run against a sandbox bucket. Your project
names, bucket and hashes will differ.

## Where you left off

The previous article left you with three things: a `store` pointing at your
bucket, a `conn` to the repo that onboards data, and `dev_dir`, its working
directory.

From here on there are **two** connections, so this article renames that one to
say which it is:

```{r}
imported <- conn   # the connection to the onboarding repo, from start-on-s3
```

Everything called `imported` below is that connection.

## The problem

You built a result from a handful of tables. Months later, half of those tables
have newer versions, because more data arrived. Someone asks which data produced
the result.

You can answer it if you wrote the versions down somewhere. Usually nobody did,
or they wrote them in a slide, or the script says `read_latest()` and the answer
no longer exists.

## What a set is

A **set** is a citable list of exact data versions. It has a name and a version
of its own, so you can refer to the whole collection with one string and get
back exactly those tables at exactly those versions -- next month, next year.

Three things it is not:

- It is not a copy of the data. It stores pointers, so making one is cheap and
  costs no storage.
- It never drifts. A set pins versions. It does not mean "whatever is current".
- It does not change the tables it points at. Writing a set is read-only on its
  members.

## Two repositories, and why

The repo from the previous article **onboards** data: files land in
`input_files/` and `datom_sync()` turns them into versioned tables. That repo
will not hold a set -- it imports data rather than building anything from it, and
datom refuses to mix the two.

So a product gets its own repo, alongside the first one in the same bucket:

```
s3://study-001-datom/
    imported/datom/     the onboarded tables      (the previous article)
    adam/datom/         the set and derived work  (this article)
```

Both prefixes are named. `imported` is the word datom uses for a table that came
from a file, as opposed to `derived` for one written from R, so the folder and
the metadata agree. `raw/`, `edc/` or `sdtm/` work just as well -- the prefix is
a plain string, and this is a convention rather than a rule.

Create it the same way as the first repo, with two extra arguments:

```{r}
# A different working directory from the onboarding repo's dev_dir.
product_dir <- fs::path(tempdir(), "study_001_adam")

# Same bucket as before, a prefix beside the onboarded data.
product_data <- datom_store_s3(
  bucket     = "study-001-datom",
  prefix     = "adam/",
  region     = "us-east-1",
  access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
  secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY")
)

product_store <- datom_store(
  data       = product_data,
  github_pat = Sys.getenv("GITHUB_PAT")
)

datom_init_repo(
  path         = product_dir,
  project_name = "STUDY_ADAM",
  store        = product_store,
  create_repo  = TRUE,
  repo_name    = "study-001-adam",
  mode         = "product",          # this repo builds; it does not onboard
  set          = "trial_product"     # the name of the set it owns
)

product <- datom_get_conn(path = product_dir, store = product_store)
print(product)
#> -- datom connection
#> * Project: "STUDY_ADAM"
#> * Backend: "s3"
#> * Role: "developer"
#> * Data root: "datom-test"
#> * Data prefix: "sets-e2e/adam/"
#> * Data region: "us-east-1"
#> * Governance: not attached
#> * Path: '/Users/.../datom-sets-e2e-product'
#> * Data repo: <https://github.com/.../datom-sets-e2e-product.git>
```

One repo owns one set, and it names that set in its own config. You do not pass
the name again when you write.

## Build the set

Look at what the onboarding repo holds:

```{r}
datom_list(imported)
#>   name  kind current_version current_data_sha         last_updated
#> 1   dm table        dc6ec312         0d29a580 2026-09-25T06:40:36Z
#> 2   ex table        e79ce54d         f261dbde 2026-09-25T06:40:16Z
```

Then declare each member and write the set. A member is one entry in the list: a
table, at one version, with optional labels.

```{r}
dm_v <- datom_history(imported, "dm")$version[1]
ex_v <- datom_history(imported, "ex")$version[1]

members <- list(
  datom_member(imported, "dm", dm_v, tags = list(type = "input", domain = "safety")),
  datom_member(imported, "ex", ex_v, tags = list(type = "input", domain = "exposure"))
)

datom_write_set(
  product,
  members,
  tags = list(description = "Trial data cut, month 1")
)
#> v Wrote set "trial_product" (2 members): "ae193a04"
```

`imported` is the connection to the onboarding repo. That is how a set cites
another project's tables: you read the version through that project's
connection, and the pointer records which project it came from.

**Labels are how you find members later.** `type` and `domain` above are your
own words -- datom attaches no meaning to them. They are worth adding, because
selecting members by label is what the editing functions later use.

If you would rather build the list step by step, for instance in a loop over
inputs you discover as you go:

```{r}
datom_assemble_set(product, tags = list(description = "Trial data cut, month 1")) |>
  datom_add_member("dm", dm_v, tags = list(type = "input", domain = "safety")) |>
  datom_add_member("ex", ex_v, tags = list(type = "input", domain = "exposure")) |>
  datom_write_set()
```

Both produce the same set. The difference is where an error appears: building
step by step, a bad version is reported on the line that added it.

## Cite it, and read it somewhere else

The set has a version, and that version is the citation:

```{r}
datom_history(product, "trial_product")[, c("version", "commit_message")]
```

Anyone with read access to the product's storage can resolve it. **No git clone
and no GitHub token are needed** -- this is the point of a set:

```{r}
reader_store <- datom_store(
  data = datom_store_s3(
    bucket     = "study-001-datom",
    prefix     = "adam/",
    region     = "us-east-1",
    access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
    secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY")
  )
  # no github_pat: that absence is what makes this a reader
)
reader <- datom_get_conn(store = reader_store, project_name = "STUDY_ADAM")

x <- datom_get_set(reader, "trial_product", version = "ae193a04")
print(x)
#> -- datom set: "trial_product"
#> * Project: "STUDY_ADAM"
#> * Version: "ae193a04aff8b757ef98fcbef228cca690b0c7332dfdd31c60e91f0680a604b5"
#> * Members: 2
#> * Tags: description=Trial data cut, month 1
#>   * dm (table) domain=safety, type=input
#>   * ex (table) domain=exposure, type=input
#> i Fetch a member with `datom_fetch_member(conn, x, "dm")`.
```

An 8-character prefix went in and the full version came back, so what you cite
is unambiguous.

To see the members as a table, one row per label:

```{r}
datom_list_members(x)
#>   name   project          version  kind    key    value
#> 1   dm STUDY_001 dc6ec312095c... table domain   safety
#> 2   dm STUDY_001 dc6ec312095c... table   type    input
```

Or grouped by a label, which is where the labels earn their place:

```{r}
datom_structure_members(x, by = "type")
#> $input
#> $input$dm
#> ...
```

Reading the set gave you pointers, not data. To get the data, ask for one member
-- through a connection to **that member's** project:

```{r}
head(datom_fetch_member(imported, x, "dm"))
#> # A tibble: 6 x 12
#>   STUDYID   DOMAIN USUBJID         SUBJID   AGE AGEU  SEX   ...
#> 1 STUDY-001 DM     STUDY-001-001        1    71 YEARS F
#> 2 STUDY-001 DM     STUDY-001-002        2    27 YEARS M
#> ...
```

This is a feature rather than an inconvenience. Reading the set needs access to
the product only, so a collection of fifty tables can be citable by someone
entitled to none of them.

## An input moves

More data arrives, and the onboarding repo takes it:

```{r}
datom_sync(imported, datom_sync_manifest(imported))
#> i Scanned 2 files: 0 new, 1 changed, 1 unchanged.
#> i Syncing 1 table...
#> v Wrote "dm" (full): "dfd8e34c"
#> v "dm" synced (changed).
#> i Sync complete: 1 succeeded, 0 failed, 1 skipped.
```

Your set still points at the old version, which is correct -- a citation does not
move on its own. When you want it to move, ask:

```{r}
x <- datom_get_set(product, "trial_product")
x <- datom_update_members(x, imported)
#> v Repointed 1 member, of 1 selected.
#> project STUDY_001:
#>   dm  9a256607 -> dfd8e34c
#> i Nothing has been written. Write the set with `datom_write_set(conn, x)`.
```

Two things worth noticing. It tells you what moved, `old -> new`, **before**
anything is written -- you are holding an edited set, and the write is yours to
make. And labels came along with the pointer; you did not have to reattach them.

```{r}
datom_write_set(product, x)
#> v Wrote set "trial_product" (1 member): "a206431c"
```

Run the refresh again when nothing has changed and it costs nothing:

```{r}
datom_write_set(product, datom_update_members(datom_get_set(product, "trial_product"),
                                              imported))
#> i No member moved: all 1 selected is already at the version this would pin.
#> i No changes detected for set "trial_product". Skipping write.
```

No new version. A set's version moves when its content moves, and not otherwise,
so re-running a build script is free.

## Retire a member

```{r}
x <- datom_get_set(product, "trial_product")
x <- datom_remove_members(x, tags = list(type = "input", domain = "exposure"))
datom_write_set(product, x)
```

Select by label rather than by position. Positions shift when somebody adds a
member; labels do not.

Removing takes no connection at all -- dropping an entry only needs the list you
already hold.

**Edits made together are written together.** If you repoint and then remove
before writing, the commit message says so:

```{r}
datom_history(product, "trial_product")[, c("version", "commit_message", "commit_sha")]
#>    version                    commit_message                                          commit_sha
#> 1 31d711b2  Update trial_product: repoint 1 member, drop 1 member   7b35a82cf101acad08c9bcd875e76b0aa7a21e3f
#> 2 ae193a04  Update trial_product                                   623d71bcddc2d512ea7cbf96cc20919a388d0cd9
```

`commit_sha` is the commit that produced each version, worked out by datom rather
than recorded by you. Note it appears here and not in your clone's copy of the
history: that file is committed *inside* the commit that would name it, so only
the copy in storage can carry it.

## Two things that surprise people

**A set can cite another set, and reading it does not open it.** You get a
pointer back, and resolving it is a separate step you choose to take. So reading
a set costs the same whether there is one table underneath or a hundred.

**A name is not a unique key.** The same table at two versions is two perfectly
valid members -- a live cut beside a frozen baseline is the usual reason. So this
returns two rows, not one:

```{r}
rows <- datom_list_members(x)
unique(rows$version[rows$name == "dm"])
```

Narrow with a label instead, or pass `tags =` to `datom_fetch_member()`, which
refuses an ambiguous name and lists the candidates rather than guessing.

## Where you are

You have a study's data onboarded in one repo, and a product in another that
cites exact versions of it. The product's version is a single string you can put
in a report, and it resolves to the same tables forever.

This scales to a product pooling several studies from several buckets; that needs
credentials with access to all of them and is a separate article.

## Teardown

If you were following along on real infrastructure, remove both repos and both
prefixes when you are done:

```{r}
datom_repo_delete(product, confirm = "STUDY_ADAM")
datom_repo_delete(imported, confirm = "STUDY_001")
```

The bucket itself is yours; datom only removes the namespaces it created.
