---
title: "Citing a Set of Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citing a Set of Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{=html}
<link rel="stylesheet" href="../inst/vignette-setup/datom.css">
```

This article continues [Starting on S3](start-on-s3.html) **in the same R
session**. It reuses that article's secrets, settings and connections, and the
four tables it onboarded into `s3://study001/imported/`.

## The aim

You want to track liver safety for study001 as data arrives. For that you keep
a small, versioned collection: the input tables you used, and the output tables
you derived from them. Any result can then be traced back to the exact data
behind it.

<div class="datom-callout datom-callout-tip" markdown="1">
**What a datom set is**

- A named, versioned list of pointers to exact table versions. It holds no data.
- It never drifts. Each member stays pinned until you move it.
- Any change, to members or to labels, makes a new version. Old versions stay
  readable.
- Reading a set needs access to its own project only. Reading a member needs
  access to that member's project.
</div>

## Set up the project

The set lives in its own project, in a folder beside the imported tables:

```
s3://study001/
    imported/datom/        onboarded tables        (Starting on S3)
    liver-safety/datom/    the set and its outputs  (this article)
```

The settings this article adds:

```{r settings-liver-safety}
project_liver_safety <- "study001-liver-safety"
prefix_liver_safety  <- "liver-safety/"
repo_liver_safety    <- "study001-liver-safety"   # GitHub repo name
set_liver_safety     <- "study001-liver-safety"   # the set this project owns
workdir_liver_safety <- fs::path(tempdir(), "study001-liver-safety")
```

A writer store and a reader store, built the same way as before:

```{r stores-liver-safety}
store_write_liver_safety <- datom_store(
  data = datom_store_s3(
    bucket     = bucket,
    prefix     = prefix_liver_safety,
    region     = region,
    access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
    secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY")
  ),
  github_pat = Sys.getenv("GITHUB_PAT")
)

store_read_liver_safety <- datom_store(
  data = datom_store_s3(
    bucket     = bucket,
    prefix     = prefix_liver_safety,
    region     = region,
    access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
    secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY")
  )
)
```

Create the repository:

```{r init-liver-safety}
datom_init_repo(
  path         = workdir_liver_safety,
  project_name = project_liver_safety,
  store        = store_write_liver_safety,
  create_repo  = TRUE,
  repo_name    = repo_liver_safety,
  mode         = "product",
  set          = set_liver_safety
)
#> v Created GitHub repo ".../study001-liver-safety".
#> v Initialized datom repository "study001-liver-safety" at '.../study001-liver-safety'
```

`mode = "product"` makes this a repository that builds from other projects'
tables instead of taking files in: it refuses `datom_sync()`, and it owns
exactly one set, the one named here. `mode` accepts `"product"` or nothing, so
`mode = "yolo"` is an error.

Connect, once to write and once to read:

```{r conns-liver-safety}
conn_write_liver_safety <- datom_get_conn(
  path  = workdir_liver_safety,
  store = store_write_liver_safety
)

conn_read_liver_safety <- datom_get_conn(
  store        = store_read_liver_safety,
  project_name = project_liver_safety
)
```

From here on, writes go through a writer connection and reads through a reader
connection.

## Version 1: register every input

Register every table the imported project holds, at its current version. Some
may go unused; that is fine. A set records exact versions, so there is nothing
to pick.

```{r v1}
tables_imported <- datom_list(conn = conn_read_imported, short_hash = FALSE)
tables_imported[, c("name", "current_version")]
#>   name                                                  current_version
#> 1   dm 773e6862cadac8e2dbc4f17927bffc21564693b420eb979c9ed80c15ea84565f
#> 2   ae 075773e9d5b3d9694e4208a5efdaa58194698f7a44d43f3eaf483077e5a70348
#> 3   ex 8dbcc9a7632a5072c5c28e8b0cc8ffd0d130f8a96e1af0f1e14673dbd0bec00b
#> 4   lb 435bccb00a61bd6e88576835d4d06740b507154d517ec1460db7994f4b20c498

inputs <- lapply(
  X   = seq_len(nrow(tables_imported)),
  FUN = function(i) {
    datom_member(
      conn    = conn_read_imported,
      name    = tables_imported$name[i],
      version = tables_imported$current_version[i],
      tags    = list(type = "input")
    )
  }
)

v1 <- datom_write_set(
  conn    = conn_write_liver_safety,
  members = inputs,
  tags    = list(description = "Liver safety, study001")
)
#> v Wrote set "study001-liver-safety" (4 members): "d2d0456b"
```

`datom_member()` takes a connection to the **member's** project, so the pointer
records which project the table belongs to. `tags` are your own labels; you will
use `type` shortly.

A set of inputs only is a valid first version. The write returns it invisibly:

```{r v1-version}
v1$metadata_sha
#> [1] "d2d0456b719244fe18d01f74a59b4e77637108405b716037c8e8b960dc15d786"
```

That string is the citation. Put it in a report and it resolves to these four
tables at these versions.

## Version 2: derive the output

The derivation is one function. Save it as `R/derive_liver_flags.R` inside
`workdir_liver_safety`:

```{r derive-script}
# R/derive_liver_flags.R
#
# One row per subject: age, sex, dose, and peak ALT and AST as a multiple of
# the upper limit of normal. Reads its inputs through the set `x`, so it uses
# exactly the versions the set pins, and records them as parents.
derive_liver_flags <- function(x, conn_input, conn_output) {
  inputs <- c("dm", "ex", "lb")

  dm <- datom_fetch_member(conn = conn_input, x = x, member = "dm")
  ex <- datom_fetch_member(conn = conn_input, x = x, member = "ex")
  lb <- datom_fetch_member(conn = conn_input, x = x, member = "lb")

  peak_xuln <- function(test) {
    rows <- lb[lb$LBTESTCD == test, ]
    tapply(X = rows$LBORRES / rows$LBORNRHI, INDEX = rows$USUBJID, FUN = max)
  }

  liver_flags <- merge(
    x  = dm[, c("USUBJID", "AGE", "SEX")],
    y  = ex[, c("USUBJID", "EXDOSE")],
    by = "USUBJID"
  )
  liver_flags$ALT_PEAK_XULN <- as.numeric(peak_xuln("ALT")[liver_flags$USUBJID])
  liver_flags$AST_PEAK_XULN <- as.numeric(peak_xuln("AST")[liver_flags$USUBJID])
  liver_flags$ELEVATED <- liver_flags$ALT_PEAK_XULN > 1 |
    liver_flags$AST_PEAK_XULN > 1

  # One parent per input used, at the version the set pins.
  pinned <- datom_list_members(x = x)
  parents <- lapply(
    X   = inputs,
    FUN = function(name) {
      datom_parent(
        conn    = conn_input,
        table   = name,
        version = unique(pinned$version[pinned$name == name])
      )
    }
  )

  datom_write(
    conn    = conn_output,
    data    = liver_flags,
    name    = "liver_flags",
    parents = parents
  )
}
```

It reads `dm`, `ex` and `lb` **through the set**, not by reaching into the
imported project for whatever is current. `ae` stays registered and unused: a
derivation can use a subset of the inputs.

Load it, run it against version 1, and add the output to the set:

```{r v2}
source(fs::path(workdir_liver_safety, "R", "derive_liver_flags.R"))

x <- datom_get_set(conn = conn_read_liver_safety, name = set_liver_safety)

liver_flags_written <- derive_liver_flags(
  x           = x,
  conn_input  = conn_read_imported,
  conn_output = conn_write_liver_safety
)
#> v Wrote "liver_flags" (full): "dafdb954"

x$members <- c(
  x$members,
  list(datom_member(
    conn    = conn_read_liver_safety,
    name    = "liver_flags",
    version = liver_flags_written$metadata_sha,
    tags    = list(type = "output")
  ))
)

v2 <- datom_write_set(
  conn          = conn_write_liver_safety,
  members       = x,
  include_paths = "R"
)
#> v Wrote set "study001-liver-safety" (5 members): "66d721f3"
```

`include_paths = "R"` commits the derivation script in the same commit as the
set, so checking out that commit gives you the pointers and the code that
produced the output. Only the derivation code is kept. The setup and onboarding
code in these articles is not, because datom already records what it did.

Every output is derived from members of the set, and records them as its
parents. So the set carries its full provenance.

## Use it

The labels you gave the members become the structure you read them through:

```{r use-structure}
x <- datom_get_set(conn = conn_read_liver_safety, name = set_liver_safety)
print(x)
#> 
#> -- datom set: "study001-liver-safety"
#> * Project: "study001-liver-safety"
#> * Version: "66d721f34ab2d8419b952fc267560218b06df6ea065669206277c1a1aa0a21ad"
#> * Members: 5
#> * Tags: description=Liver safety, study001
#>   * ae (table) type=input
#>   * dm (table) type=input
#>   * ex (table) type=input
#>   * lb (table) type=input
#>   * liver_flags (table) type=output
#> i Fetch a member with `datom_fetch_member(conn, x, "ae")`.

dp <- datom_structure_members(x = x, by = "type")

head(dp$output$liver_flags(conn = conn_read_liver_safety))
#> # A tibble: 6 x 7
#>   USUBJID         AGE SEX   EXDOSE ALT_PEAK_XULN AST_PEAK_XULN ELEVATED
#>   <chr>         <int> <chr>  <int>         <dbl>         <dbl> <lgl>   
#> 1 STUDY-001-001    71 F        200         0.416         0.648 FALSE   
#> 2 STUDY-001-002    27 M        200         0.711         0.585 FALSE   
#> 3 STUDY-001-003    35 M          0         0.911         0.455 FALSE   
#> 4 STUDY-001-004    68 M        200         0.752         0.728 FALSE   
#> 5 STUDY-001-005    43 M        200         0.846         0.535 FALSE   
#> 6 STUDY-001-006    60 F          0         0.404         0.722 FALSE

nrow(dp$input$lb(conn = conn_read_imported))
#> [1] 205
```

`type = "input"` and `type = "output"` are what produce `dp$input` and
`dp$output`. Each entry is a function: call it with a connection to the
member's project and it reads that member at its pinned version.

The same members as a data frame, one row per label. It also carries each
member's full `version` and `kind`, left out here to fit the page:

```{r use-list}
datom_list_members(x = x)[, c("name", "project", "key", "value")]
#>          name               project  key  value
#> 1          ae     study001-imported type  input
#> 2          dm     study001-imported type  input
#> 3          ex     study001-imported type  input
#> 4          lb     study001-imported type  input
#> 5 liver_flags study001-liver-safety type output
```

The set's own history, and version 1 read back by its version:

```{r use-history}
set_history <- datom_history(conn = conn_read_liver_safety,
                             name = set_liver_safety, short_hash = TRUE)
set_history[, c("version", "commit_message")]
#>    version               commit_message
#> 1 66d721f3 Update study001-liver-safety
#> 2 d2d0456b Update study001-liver-safety

datom_get_set(
  conn    = conn_read_liver_safety,
  name    = set_liver_safety,
  version = v1$metadata_sha
)
#> 
#> -- datom set: "study001-liver-safety"
#> * Project: "study001-liver-safety"
#> * Version: "d2d0456b719244fe18d01f74a59b4e77637108405b716037c8e8b960dc15d786"
#> * Members: 4
#> * Tags: description=Liver safety, study001
#>   * ae (table) type=input
#>   * dm (table) type=input
#>   * ex (table) type=input
#>   * lb (table) type=input
#> i Fetch a member with `datom_fetch_member(conn, x, "ae")`.
```

## Refresh: new data arrives

The month-4 extract arrives for all four tables. The imported project takes it
in, as before:

```{r refresh-sync}
for (domain in c("dm", "ex", "lb", "ae")) {
  write.csv(
    x         = datom_example_data(domain = domain, cutoff_date = "2026-04-28"),
    file      = fs::path(inputs_imported, paste0(domain, ".csv")),
    row.names = FALSE
  )
}

manifest <- datom_sync_manifest(conn = conn_write_imported)
#> i Scanned 4 files: 0 new, 4 changed, 0 unchanged.

synced <- datom_sync(conn = conn_write_imported, manifest = manifest)
#> i Syncing 4 tables...
#> v Wrote "ae" (full): "97e2a95a"
#> v "ae" synced (changed).
#> v Wrote "dm" (full): "a87789e0"
#> v "dm" synced (changed).
#> v Wrote "ex" (full): "b59939d6"
#> v "ex" synced (changed).
#> v Wrote "lb" (full): "dc682195"
#> v "lb" synced (changed).
#> i Sync complete: 4 succeeded, 0 failed, 0 skipped.
```

The set still points at the month-3 versions. A citation does not move on its
own. Move it in this order: the inputs, then the output derived from them, then
one write.

```{r refresh-update}
x <- datom_get_set(conn = conn_read_liver_safety, name = set_liver_safety)

x <- datom_update_members(
  x    = x,
  conn = conn_read_imported,
  tags = list(type = "input")
)
#> v Repointed 4 members, of 4 selected.
#> project study001-imported:
#>   ae  075773e9 -> 97e2a95a
#>   dm  773e6862 -> a87789e0
#>   ex  8dbcc9a7 -> b59939d6
#>   lb  435bccb0 -> dc682195
#> i Nothing has been written. Write the set with `datom_write_set(conn, x)`.

derive_liver_flags(
  x           = x,
  conn_input  = conn_read_imported,
  conn_output = conn_write_liver_safety
)
#> v Wrote "liver_flags" (full): "0f89dd1b"

x <- datom_update_members(
  x    = x,
  conn = conn_read_liver_safety,
  tags = list(type = "output")
)
#> v Repointed 1 member, of 1 selected.
#> project study001-liver-safety:
#>   liver_flags  dafdb954 -> 0f89dd1b
#> i Nothing has been written. Write the set with `datom_write_set(conn, x)`.

v3 <- datom_write_set(
  conn          = conn_write_liver_safety,
  members       = x,
  include_paths = "R"
)
#> v Wrote set "study001-liver-safety" (5 members): "56013493"
```

`datom_update_members()` moves the selected members to their project's current
version and reports what moved. Nothing is written until `datom_write_set()`.

The derivation runs on `x` as you hold it, with the inputs already moved, so the
new `liver_flags` is built from month-4 data. datom does not check the order for
you. Writing once, after both the inputs and the output have moved, is what
keeps version 3 from pairing new inputs with an output built from old ones.

## Who can read what

A set and one reader connection per project open everything in it:

| You hold | You can read |
|---|---|
| `conn_read_liver_safety` | the set, every version of it, and `liver_flags` |
| `conn_read_imported` | the input tables |

Neither needs a GitHub token or a clone. Someone who can read the liver-safety
project but not the imported one can still read and cite the set, and see which
inputs it names; they cannot open the inputs.

## Where you are

One project holds the study's onboarded tables. A second holds a set that names
exact versions of them, plus the output derived from them, the code that derived
it, and three versions of history. One string cites any of those versions.

Pooling several studies into one set is a separate article.

## Teardown

Delete both projects, storage first and then the repository each time:

```{r teardown}
datom_storage_delete_prefix(conn = conn_write_liver_safety)
datom_repo_delete(conn = conn_write_liver_safety, confirm = project_liver_safety)

datom_storage_delete_prefix(conn = conn_write_imported)
datom_repo_delete(conn = conn_write_imported, confirm = project_imported)
```

`datom_storage_delete_prefix()` deletes everything under that project's
`datom/` folder, and it does not ask first. The rest of the bucket, and the
bucket itself, are left alone. `datom_repo_delete()` deletes the GitHub
repository and the local clone, and does not touch storage.
