---
title: "Clinical Data Versioning with tbit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical Data Versioning with tbit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",

eval = FALSE
)
```

## The Problem

Clinical trials generate evolving data. As subjects enroll and visits accumulate, EDC (Electronic Data Capture) systems produce periodic data extracts—snapshots that grow over time. A typical Phase II study might have 50–100 tables updated monthly, and your analysis code must work reproducibly against any of those snapshots.

tbit solves this by version-controlling your data alongside code. Metadata lives in git; actual data lives in S3 as content-addressed parquet files. Every extract gets a SHA-based version, so you (or a colleague, or a regulator) can reproduce any historical result.

This vignette walks through a realistic workflow: receiving monthly EDC extracts for a Phase II study, syncing them to tbit, deriving an analysis dataset, and accessing historical versions.

## Example Study: STUDY-001

tbit ships with simulated data for a small Phase II study:

- **48 subjects** enrolled over 6 months (Jan–Jun 2026)
- **DM domain**: demographics (age, sex, race, country, enrollment date)
- **EX domain**: exposure (treatment arm, dose, first dose date)
- **Monthly snapshots**: filtering by enrollment date simulates receiving updated EDC extracts

```{r}
library(tbit)

# View the monthly cutoff dates
tbit_example_cutoffs()
#> month_1    month_2    month_3    month_4    month_5    month_6
#> "2026-01-28" "2026-02-28" "2026-03-28" "2026-04-28" "2026-05-28" "2026-06-28"

# Month-1 extract: only early enrollees
dm_m1 <- tbit_example_data("dm", cutoff_date = "2026-01-28")
nrow(dm_m1)
#> [1] 6

# Month-3 extract: enrollment ramps up
dm_m3 <- tbit_example_data("dm", cutoff_date = "2026-03-28")
nrow(dm_m3)
#> [1] 19

# Full study data
dm_all <- tbit_example_data("dm")
nrow(dm_all)
#> [1] 48
```

## Setup

### One-Time Repository Initialization

A data developer sets up the tbit repository once. This creates the folder structure, initializes git with a remote, and generates configuration files.

```{r}
# Required environment variables (set these in .Renviron or your CI environment)
# TBIT_STUDY_001_ACCESS_KEY_ID = "your_aws_key"
# TBIT_STUDY_001_SECRET_ACCESS_KEY = "your_aws_secret"
# GITHUB_PAT = "your_github_pat"

tbit_init_repo(
  path = "study_001_data",
  project_name = "STUDY_001",
  remote_url = "https://github.com/org/study-001-data.git",
  bucket = "clinical-data-bucket",
  prefix = "study-001/"
)
```

This creates:

```
study_001_data/
├── .tbit/
│   ├── project.yaml       # S3 bucket, credentials config
│   ├── routing.json        # Read method dispatch
│   └── manifest.json       # Table catalog (initially empty)
├── input_files/            # Drop EDC extracts here (gitignored)
└── .gitignore              # Keeps data files out of git
```

### Connect

```{r}
conn <- tbit_get_conn(path = "study_001_data")
print(conn)
#> ── tbit connection ──
#> • Project: STUDY_001
#> • Role: developer
#> • Bucket: clinical-data-bucket
#> • Prefix: study-001/
#> • Region: us-east-1
#> • Path: /path/to/study_001_data
```

## Monthly Workflow: Receiving and Syncing EDC Extracts

### Month 1: First Data Transfer

Your data management team delivers the first EDC extract. You place the files in `input_files/`:

```{r}
# Simulate receiving the month-1 extract
cutoffs <- tbit_example_cutoffs()

dm_m1 <- tbit_example_data("dm", cutoff_date = cutoffs["month_1"])
ex_m1 <- tbit_example_data("ex", cutoff_date = cutoffs["month_1"])

write.csv(dm_m1, "study_001_data/input_files/dm.csv", row.names = FALSE)
write.csv(ex_m1, "study_001_data/input_files/ex.csv", row.names = FALSE)
```

Scan and sync:

```{r}
# Scan input_files/ — computes SHA of each file, compares against manifest
manifest <- tbit_sync_manifest(conn)
#> ℹ Scanned 2 files: 2 new, 0 changed, 0 unchanged.
manifest
#>   name          file format  file_sha                           status
#> 1   dm .../dm.csv    csv    a1b2c3...                          new
#> 2   ex .../ex.csv    csv    d4e5f6...                          new

# Sync: imports each file, converts to parquet, uploads to S3, commits to git
results <- tbit_sync(conn, manifest)
#> ℹ Syncing 2 tables...
#> ✔ dm synced (new).
#> ✔ ex synced (new).
#> ℹ Sync complete: 2 succeeded, 0 failed, 0 skipped.
```

What just happened:

1. Each CSV was imported via `rio::import()` and written as parquet to S3
2. A SHA-256 was computed for both the source file and the parquet data
3. `metadata.json` and `version_history.json` were created in git and synced to S3
4. A git commit was made and pushed for each table
5. The local `manifest.json` was updated with current version info

### Deriving ADSL

Now we create a subject-level analysis dataset by joining DM and EX:

```{r}
# Read the versioned data back from tbit
dm <- tbit_read(conn, "dm")
ex <- tbit_read(conn, "ex")

# Derive ADSL: demographics + treatment info
adsl <- merge(
  dm[, c("USUBJID", "SUBJID", "AGE", "SEX", "RACE", "COUNTRY", "RFSTDTC")],
  ex[, c("USUBJID", "EXTRT", "EXDOSE", "EXSTDTC")],
  by = "USUBJID"
)

# Add derived fields
adsl$AGEGR1 <- cut(
  adsl$AGE,
  breaks = c(0, 40, 60, Inf),
  labels = c("<40", "40-60", ">60"),
  right = FALSE
)
adsl$TRT01A <- adsl$EXTRT
adsl$TRTSDT <- adsl$EXSTDTC

nrow(adsl)
#> [1] 6
table(adsl$TRT01A)
#> DRUG-X 200mg      PLACEBO
#>            3            3

# Write derived ADSL to tbit
tbit_write(conn, data = adsl, name = "ADSL", message = "Derive ADSL from month-1 data")
#> ✔ Wrote "ADSL" (full): "f7a8b9c2"
```

### Month 3: Updated Extract

Two months later, enrollment has progressed. The data management team sends updated extracts:

```{r}
# Simulate month-3 extract
dm_m3 <- tbit_example_data("dm", cutoff_date = cutoffs["month_3"])
ex_m3 <- tbit_example_data("ex", cutoff_date = cutoffs["month_3"])

write.csv(dm_m3, "study_001_data/input_files/dm.csv", row.names = FALSE)
write.csv(ex_m3, "study_001_data/input_files/ex.csv", row.names = FALSE)

# Scan — tbit detects both files have changed (different SHA)
manifest <- tbit_sync_manifest(conn)
#> ℹ Scanned 2 files: 0 new, 2 changed, 0 unchanged.

# Sync the updates
results <- tbit_sync(conn, manifest)
#> ℹ Syncing 2 tables...
#> ✔ dm synced (changed).
#> ✔ ex synced (changed).
#> ℹ Sync complete: 2 succeeded, 0 failed, 0 skipped.
```

Because the parquet content changed (new subjects), tbit uploaded new parquet files to S3 with fresh SHAs. The old parquet files remain — they are never overwritten or deleted.

```{r}
# Re-derive ADSL with updated data
dm <- tbit_read(conn, "dm")
ex <- tbit_read(conn, "ex")

adsl <- merge(
  dm[, c("USUBJID", "SUBJID", "AGE", "SEX", "RACE", "COUNTRY", "RFSTDTC")],
  ex[, c("USUBJID", "EXTRT", "EXDOSE", "EXSTDTC")],
  by = "USUBJID"
)
adsl$AGEGR1 <- cut(adsl$AGE, breaks = c(0, 40, 60, Inf),
                    labels = c("<40", "40-60", ">60"), right = FALSE)
adsl$TRT01A <- adsl$EXTRT
adsl$TRTSDT <- adsl$EXSTDTC

nrow(adsl)
#> [1] 19
table(adsl$TRT01A)
#> DRUG-X 200mg      PLACEBO
#>           10            9

tbit_write(conn, data = adsl, name = "ADSL", message = "Update ADSL with month-3 data")
#> ✔ Wrote "ADSL" (full): "c3d4e5f6"
```

## Querying Version History

### List Tables

```{r}
tbit_list(conn, include_versions = TRUE)
#>   name current_version current_data_sha last_updated          version_count
#> 1 ADSL c3d4e5f6...     b2c3d4...        2026-03-28T10:30:00Z  2
#> 2 dm   a8b9c0d1...     e5f6a7...        2026-03-28T10:25:00Z  2
#> 3 ex   d1e2f3a4...     f6a7b8...        2026-03-28T10:25:00Z  2
```

### View ADSL Version History

```{r}
tbit_history(conn, "ADSL")
#>   version      data_sha     timestamp             author          commit_message
#> 1 c3d4e5f6...  b2c3d4...    2026-03-28T10:30:00Z  jane@co.com    Update ADSL with month-3 data
#> 2 f7a8b9c2...  a1b2c3...    2026-01-28T10:30:00Z  jane@co.com    Derive ADSL from month-1 data
```

### Reproduce a Prior Result

A colleague needs to verify the month-1 ADSL. They use the version hash from the history:

```{r}
# Read the exact month-1 version
adsl_v1 <- tbit_read(conn, "ADSL", version = "f7a8b9c2...")
nrow(adsl_v1)
#> [1] 6

# Same result as month-1 derivation — reproducible
```

## Data Reader Workflow

Downstream analysts don't need git access. They connect directly to S3:

```{r}
# Data reader only needs S3 credentials
# TBIT_STUDY_001_ACCESS_KEY_ID and TBIT_STUDY_001_SECRET_ACCESS_KEY

reader_conn <- tbit_get_conn(
  bucket = "clinical-data-bucket",
  prefix = "study-001/",
  project_name = "STUDY_001"
)
print(reader_conn)
#> ── tbit connection ──
#> • Project: STUDY_001
#> • Role: reader
#> • Bucket: clinical-data-bucket

# Readers can list, read, and access any version
tables <- tbit_list(reader_conn)
adsl <- tbit_read(reader_conn, "ADSL")
```

## Repository Validation

After syncing, verify git and S3 are consistent:

```{r}
tbit_validate(conn)
#> ✔ All checks passed. Git and S3 are consistent.

# Check repository status
tbit_status(conn)
#> ── tbit status ──
#> ℹ Project: "STUDY_001"
#> ℹ Bucket: "clinical-data-bucket"
#> ℹ Role: "developer"
#> ✔ Git: clean (no uncommitted changes)
#> ℹ Branch: "main"
#> ℹ Tables on S3: 3
#> ✔ Input files: all 2 unchanged
```

## Change Detection

tbit uses SHA-based content addressing for efficient change detection. If the data hasn't changed, no upload occurs:

```{r}
# Re-sync with the same files — nothing happens
manifest <- tbit_sync_manifest(conn)
#> ℹ Scanned 2 files: 0 new, 0 changed, 2 unchanged.

results <- tbit_sync(conn, manifest)
#> ℹ All files unchanged. Nothing to sync.
```

Similarly, re-writing the same ADSL data is a no-op:

```{r}
tbit_write(conn, data = adsl, name = "ADSL", message = "No change")
#> ℹ No changes detected for "ADSL". Skipping write.
```

## Key Concepts

### Content Addressing

Every data file is stored as `{sha256}.parquet`. Two identical datasets always produce the same SHA, so there's built-in deduplication. Different data produces a different SHA, so versions never collide.

### Metadata vs Data

- **Metadata** (in git): `metadata.json` tracks dimensions, column names, data SHA, and timestamps. `version_history.json` maps each tbit version to its data SHA.
- **Data** (in S3): Immutable parquet files, keyed by SHA. Never overwritten or deleted.

### tbit Version = metadata SHA

The tbit version identifier is the SHA of the metadata fields (sorted alphabetically). Since metadata includes the data SHA, the version uniquely identifies a (data, metadata) pair. Changing a custom metadata field (like a description) produces a new version even if the data is identical.

### Two User Roles

| Role | Credentials | Capabilities |
|------|-------------|--------------|
| **Developer** | AWS + GITHUB_PAT | Read, write, sync, validate |
| **Reader** | AWS only | Read any version |

Role is auto-detected: if `GITHUB_PAT` is set and a local repo path is provided, you're a developer.

## Summary

This workflow demonstrates the core tbit loop for clinical data:

1. **Receive** EDC extracts → place in `input_files/`
2. **Scan** with `tbit_sync_manifest()` → detect new and changed files
3. **Sync** with `tbit_sync()` → import, convert to parquet, version, upload
4. **Derive** analysis datasets → `tbit_read()` + transformation + `tbit_write()`
5. **Reproduce** → `tbit_read(conn, name, version = "...")` retrieves any historical version

Every step is tracked in git with full audit trail. Data readers access any version via S3 without needing git credentials.
