---
title: "Clinical Data Versioning with tbit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical Data Versioning with tbit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## The Problem

Clinical trials generate evolving data. As subjects enroll and visits accumulate, EDC (Electronic Data Capture) systems produce periodic data extracts—snapshots that grow over time. A typical Phase II study might have 50–100 tables updated monthly, and your analysis code must work reproducibly against any of those snapshots.

tbit solves this by version-controlling your raw data alongside code. Metadata lives in git; actual data lives in S3 as content-addressed parquet files. Every extract gets a SHA-based version, so you (or a colleague, or a regulator) can reproduce any historical result.

This vignette walks through a realistic workflow: receiving monthly EDC extracts for a Phase II study, syncing them to tbit, listing tables and versions, and retrieving historical snapshots by version.

> **Note**: tbit manages *raw* versioned data. Derived datasets (like ADSL) are built downstream by the daapr ecosystem, which reads versioned inputs from tbit.

## Example Study: STUDY-001

tbit ships with simulated data for a small Phase II study:

- **48 subjects** enrolled over 6 months (Jan–Jun 2026)
- **DM domain**: demographics (age, sex, race, country, enrollment date)
- **EX domain**: exposure (treatment arm, dose, first dose date)
- **Monthly snapshots**: filtering by enrollment date simulates receiving updated EDC extracts

```{r}
library(tbit)

# View the monthly cutoff dates
tbit_example_cutoffs()
#> month_1    month_2    month_3    month_4    month_5    month_6
#> "2026-01-28" "2026-02-28" "2026-03-28" "2026-04-28" "2026-05-28" "2026-06-28"

# Month-1 extract: only early enrollees
dm_m1 <- tbit_example_data("dm", cutoff_date = "2026-01-28")
nrow(dm_m1)
#> [1] 6

# Month-3 extract: enrollment ramps up
dm_m3 <- tbit_example_data("dm", cutoff_date = "2026-03-28")
nrow(dm_m3)
#> [1] 19

# Full study data
dm_all <- tbit_example_data("dm")
nrow(dm_all)
#> [1] 48
```

## Setup

### One-Time Repository Initialization

A data developer sets up the tbit repository once. This creates the folder structure, initializes git with a remote, and generates configuration files.

```{r}
# Required environment variables (set these in .Renviron or your CI environment)
# TBIT_STUDY_001_ACCESS_KEY_ID = "your_aws_key"
# TBIT_STUDY_001_SECRET_ACCESS_KEY = "your_aws_secret"
# GITHUB_PAT = "your_github_pat"

tbit_init_repo(
  path = "study_001_data",
  project_name = "STUDY_001",
  remote_url = "https://github.com/org/study-001-data.git",
  bucket = "clinical-data-bucket",
  prefix = "study-001/"
)
```

This creates:

```
study_001_data/
├── .tbit/
│   ├── project.yaml       # S3 bucket, credentials config
│   ├── routing.json        # Read method dispatch
│   └── manifest.json       # Table catalog (initially empty)
├── input_files/            # Drop EDC extracts here (gitignored)
└── .gitignore              # Keeps data files out of git
```

### Connect

```{r}
conn <- tbit_get_conn(path = "study_001_data")
print(conn)
#> ── tbit connection ──
#> • Project: STUDY_001
#> • Role: developer
#> • Bucket: clinical-data-bucket
#> • Prefix: study-001/
#> • Region: us-east-1
#> • Path: /path/to/study_001_data
```

## Monthly Workflow: Receiving and Syncing EDC Extracts

### Month 1: First Data Transfer

Your data management team delivers the first EDC extract. You place the files in `input_files/`:

```{r}
# Simulate receiving the month-1 extract
cutoffs <- tbit_example_cutoffs()

dm_m1 <- tbit_example_data("dm", cutoff_date = cutoffs["month_1"])
ex_m1 <- tbit_example_data("ex", cutoff_date = cutoffs["month_1"])

write.csv(dm_m1, "study_001_data/input_files/dm.csv", row.names = FALSE)
write.csv(ex_m1, "study_001_data/input_files/ex.csv", row.names = FALSE)
```

Scan and sync:

```{r}
# Scan input_files/ — computes SHA of each file, compares against manifest
manifest <- tbit_sync_manifest(conn)
#> ℹ Scanned 2 files: 2 new, 0 changed, 0 unchanged.
manifest
#>   name          file format  file_sha                           status
#> 1   dm .../dm.csv    csv    a1b2c3...                          new
#> 2   ex .../ex.csv    csv    d4e5f6...                          new

# Sync: imports each file, converts to parquet, uploads to S3, commits to git
results <- tbit_sync(conn, manifest)
#> ℹ Syncing 2 tables...
#> ✔ dm synced (new).
#> ✔ ex synced (new).
#> ℹ Sync complete: 2 succeeded, 0 failed, 0 skipped.
```

What just happened:

1. Each CSV was imported via `rio::import()` and written as parquet to S3
2. A SHA-256 was computed for both the source file and the parquet data
3. `metadata.json` and `version_history.json` were created in git and synced to S3
4. A git commit was made and pushed for each table
5. The local `manifest.json` was updated with current version info

## Exploring the Repository After Syncing

Before moving on, it's worth pausing to verify the state of the repository and get familiar with the data you've just versioned.

### List All Tables

`tbit_list()` gives you a quick inventory of what is registered and when each table was last updated:

```{r}
tbit_list(conn)
#>   name current_version last_updated          version_count
#> 1 dm   f7a8b9c2...     2026-01-28T10:25:00Z  1
#> 2 ex   c3d4e5f6...     2026-01-28T10:25:00Z  1
```

Pass `include_versions = TRUE` to see the current data SHA alongside the version identifier:

```{r}
tbit_list(conn, include_versions = TRUE)
#>   name current_version current_data_sha last_updated          version_count
#> 1 dm   f7a8b9c2...     a1b2c3...        2026-01-28T10:25:00Z  1
#> 2 ex   c3d4e5f6...     d4e5f6...        2026-01-28T10:25:00Z  1
```

### Inspect Version History

`tbit_history()` shows the full audit trail for a single table — who synced it, when, and which version hash corresponds to each snapshot:

```{r}
tbit_history(conn, "dm")
#>   version      data_sha     timestamp             author          commit_message
#> 1 f7a8b9c2...  a1b2c3...    2026-01-28T10:25:00Z  jane@co.com    Sync dm
```

Each version hash uniquely identifies the (data, metadata) pair at that point in time. Right now there is one entry — the month-1 snapshot.

### Pull Down a Version to Evaluate

Before trusting that the sync worked correctly, read the data back and spot-check it:

```{r}
# Read the current (month-1) snapshot
dm_m1 <- tbit_read(conn, "dm")
nrow(dm_m1)
#> [1] 6

# Spot-check: confirm only early enrollees are present
range(as.Date(dm_m1$RFSTDTC))
#> [1] "2026-01-03" "2026-01-28"
```

Rows, columns, and date ranges all look right. You can also read by the version hash directly — short prefix matching works just like `git log --abbrev-commit`:

```{r}
# Equivalent: read by version hash prefix
dm_m1 <- tbit_read(conn, "dm", version = "f7a8b9c2")
nrow(dm_m1)
#> [1] 6
```

## Month 3: Updated Extract

Two months later, enrollment has progressed. The data management team sends updated extracts:

```{r}
# Simulate month-3 extract
dm_m3 <- tbit_example_data("dm", cutoff_date = cutoffs["month_3"])
ex_m3 <- tbit_example_data("ex", cutoff_date = cutoffs["month_3"])

write.csv(dm_m3, "study_001_data/input_files/dm.csv", row.names = FALSE)
write.csv(ex_m3, "study_001_data/input_files/ex.csv", row.names = FALSE)

# Scan — tbit detects both files have changed (different SHA)
manifest <- tbit_sync_manifest(conn)
#> ℹ Scanned 2 files: 0 new, 2 changed, 0 unchanged.

# Sync the updates
results <- tbit_sync(conn, manifest)
#> ℹ Syncing 2 tables...
#> ✔ dm synced (changed).
#> ✔ ex synced (changed).
#> ℹ Sync complete: 2 succeeded, 0 failed, 0 skipped.
```

Because the parquet content changed (new subjects), tbit uploaded new parquet files to S3 with fresh SHAs. The old parquet files remain — they are never overwritten or deleted.

## Change Detection

tbit uses SHA-based content addressing for efficient change detection. If the data hasn't changed, no upload occurs:

```{r}
# Re-sync with the same files — nothing happens
manifest <- tbit_sync_manifest(conn)
#> ℹ Scanned 2 files: 0 new, 0 changed, 2 unchanged.

results <- tbit_sync(conn, manifest)
#> ℹ All files unchanged. Nothing to sync.
```

## Data Reader Workflow

Downstream analysts don't need git access. They connect directly to S3:

```{r}
# Data reader only needs S3 credentials
# TBIT_STUDY_001_ACCESS_KEY_ID and TBIT_STUDY_001_SECRET_ACCESS_KEY

reader_conn <- tbit_get_conn(
  bucket = "clinical-data-bucket",
  prefix = "study-001/",
  project_name = "STUDY_001"
)
print(reader_conn)
#> ── tbit connection ──
#> • Project: STUDY_001
#> • Role: reader
#> • Bucket: clinical-data-bucket

# Readers can list, read, and access any version
tables <- tbit_list(reader_conn)
dm <- tbit_read(reader_conn, "dm")
```

## Repository Validation

After syncing, verify git and S3 are consistent:

```{r}
tbit_validate(conn)
#> ✔ All checks passed. Git and S3 are consistent.

# Check repository status
tbit_status(conn)
#> ── tbit status ──
#> ℹ Project: "STUDY_001"
#> ℹ Bucket: "clinical-data-bucket"
#> ℹ Role: "developer"
#> ✔ Git: clean (no uncommitted changes)
#> ℹ Branch: "main"
#> ℹ Tables on S3: 2
#> ✔ Input files: all 2 unchanged
```

## Key Concepts

### Content Addressing

Every data file is stored as `{sha256}.parquet`. Two identical datasets always produce the same SHA, so there's built-in deduplication. Different data produces a different SHA, so versions never collide.

### Metadata vs Data

- **Metadata** (in git): `metadata.json` tracks dimensions, column names, data SHA, and timestamps. `version_history.json` maps each tbit version to its data SHA.
- **Data** (in S3): Immutable parquet files, keyed by SHA. Never overwritten or deleted.

### tbit Version = metadata SHA

The tbit version identifier is the SHA of the metadata fields (sorted alphabetically). Since metadata includes the data SHA, the version uniquely identifies a (data, metadata) pair. You can use a short prefix of the version hash (like `"f7a8b9"`) — tbit resolves it as long as it's unambiguous, just like git short SHAs.

### Two User Roles

| Role | Credentials | Capabilities |
|------|-------------|--------------|
| **Developer** | AWS + GITHUB_PAT | Read, write, sync, validate |
| **Reader** | AWS only | Read any version |

Role is auto-detected: if `GITHUB_PAT` is set and a local repo path is provided, you're a developer.

## Summary

This workflow demonstrates the core tbit loop for clinical data:

1. **Receive** EDC extracts → place in `input_files/`
2. **Scan** with `tbit_sync_manifest()` → detect new and changed files
3. **Sync** with `tbit_sync()` → import, convert to parquet, version, upload
4. **Explore** with `tbit_list()`, `tbit_history()`, and `tbit_read()` → verify what was synced before moving on
5. **Update** by dropping new extracts and repeating the sync cycle
6. **Retrieve** any historical snapshot with `tbit_read(conn, name, version = "...")`

Every step is tracked in git with full audit trail. Data readers access any version via S3 without needing git credentials.

Derived datasets (like ADSL) belong in the downstream daapr ecosystem, which reads versioned raw inputs managed by tbit.
