---
title: "Credential Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Credential Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

datom uses environment variables for all credentials — no passwords in code, no
configuration files committed to git. This vignette explains how credentials
work, how to set them up, and how to manage them across multiple projects.

## How datom Derives Credential Names

datom generates environment variable names deterministically from your
`project_name`. The formula is:

```
DATOM_{PROJECT_NAME}_ACCESS_KEY_ID
DATOM_{PROJECT_NAME}_SECRET_ACCESS_KEY
```

For example:

| `project_name` | Access key env var | Secret key env var |
|-----------------|--------------------|--------------------|
| `STUDY_001`     | `DATOM_STUDY_001_ACCESS_KEY_ID` | `DATOM_STUDY_001_SECRET_ACCESS_KEY` |
| `REGISTRY_DATA` | `DATOM_REGISTRY_DATA_ACCESS_KEY_ID` | `DATOM_REGISTRY_DATA_SECRET_ACCESS_KEY` |

This means:

- Every team member on the **same project** uses the **same env var names**
- Each person sets their **own values** (personal or shared IAM keys)
- Projects with **different names** never collide

## Developer vs Reader Credentials

| | Developer | Reader |
|---|-----------|--------|
| **S3 access** | Read + write | Read only |
| **Git access** | Yes (`GITHUB_PAT`) | No |
| **AWS keys** | `DATOM_{PROJECT}_ACCESS_KEY_ID` | `DATOM_{PROJECT}_ACCESS_KEY_ID` |
| | `DATOM_{PROJECT}_SECRET_ACCESS_KEY` | `DATOM_{PROJECT}_SECRET_ACCESS_KEY` |

datom auto-detects your role based on the presence of `GITHUB_PAT`:

- **`GITHUB_PAT` set** → developer (git push/pull enabled)
- **`GITHUB_PAT` missing** → reader (S3-only access)

Reader AWS keys should be scoped to read-only S3 access via IAM policies.

## Setting Credentials

### `.Renviron` (Recommended for Local Development)

Add credentials to your user-level `~/.Renviron` file (restart R after editing):

```bash
# AWS credentials for your datom project
DATOM_STUDY_001_ACCESS_KEY_ID=AKIAIOSFODNN7EXAMPLE
DATOM_STUDY_001_SECRET_ACCESS_KEY=wJalrXUtnFEMI/K7MDENG/bPxRfiCYEXAMPLEKEY

# GitHub PAT (developers only — enables git push/pull)
GITHUB_PAT=ghp_xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx
```

You can edit `.Renviron` from R with `usethis::edit_r_environ()`.

> **Security**: `~/.Renviron` is per-user and lives outside any git repository.
> Never commit credentials to version control.

### CI/CD Environments

In CI/CD (GitHub Actions, GitLab CI, etc.), set credentials as encrypted
secrets:
```yaml
# Example: GitHub Actions
env:
  DATOM_STUDY_001_ACCESS_KEY_ID: ${{ secrets.DATOM_STUDY_001_ACCESS_KEY_ID }}
  DATOM_STUDY_001_SECRET_ACCESS_KEY: ${{ secrets.DATOM_STUDY_001_SECRET_ACCESS_KEY }}
  GITHUB_PAT: ${{ secrets.GITHUB_TOKEN }}
```

### `keyring` (Optional — Encrypted Credential Store)

For enhanced security, use the
[keyring](https://r-lib.github.io/keyring/) package to store credentials
in your OS keychain and load them at session start:

```{r}
# One-time setup: store credentials
keyring::key_set("DATOM_STUDY_001_ACCESS_KEY_ID")
keyring::key_set("DATOM_STUDY_001_SECRET_ACCESS_KEY")

# In your .Rprofile or session startup script:
Sys.setenv(
  DATOM_STUDY_001_ACCESS_KEY_ID = keyring::key_get("DATOM_STUDY_001_ACCESS_KEY_ID"),
  DATOM_STUDY_001_SECRET_ACCESS_KEY = keyring::key_get("DATOM_STUDY_001_SECRET_ACCESS_KEY")
)
```

This keeps credentials encrypted at rest and avoids plain-text `.Renviron` files.

## Multiple Projects

If you work on several datom projects, each project uses its own env var pair.
There's no cross-contamination:

```bash
# ~/.Renviron

# Project 1
DATOM_STUDY_001_ACCESS_KEY_ID=AKIA...
DATOM_STUDY_001_SECRET_ACCESS_KEY=...

# Project 2
DATOM_REGISTRY_DATA_ACCESS_KEY_ID=AKIA...
DATOM_REGISTRY_DATA_SECRET_ACCESS_KEY=...

# Shared across all projects (developer role)
GITHUB_PAT=ghp_...
```

datom only reads the env vars matching the `project_name` in
`.datom/project.yaml` (or the `project_name` you pass to `datom_get_conn()`).
Credentials for unrelated projects are never touched.

## What to Share with Teammates

### For Developers

Share the project setup instructions (typically in the repository README):

| Item | Example |
|------|---------|
| Git remote URL | `https://github.com/org/study-001-data.git` |
| AWS access key | Distribute via your team's secrets manager |
| AWS secret key | Distribute via your team's secrets manager |
| `GITHUB_PAT` | Each developer generates their own via [GitHub Settings](https://github.com/settings/tokens) |

### For Readers

Readers don't need git access. Share:

| Item | Example |
|------|---------|
| `project_name` | `"STUDY_001"` |
| `bucket` | `"clinical-data-bucket"` |
| `prefix` | `"study-001/"` |
| AWS credentials | Reader-scoped IAM keys (read-only S3 access) |

Readers connect with:

```{r}
conn <- datom_get_conn(
  project_name = "STUDY_001",
  bucket = "clinical-data-bucket",
  prefix = "study-001/"
)
```

## Troubleshooting

### "Missing credential" error

```
✖ Missing environment variable: DATOM_STUDY_001_ACCESS_KEY_ID
ℹ Set it in ~/.Renviron or your environment.
```

Check that the env var is set and the `project_name` matches exactly (case-sensitive).

### Developer detected as reader

If `datom_get_conn()` returns `role = "reader"` when you expect developer:

```{r}
# Check if GITHUB_PAT is visible
Sys.getenv("GITHUB_PAT")
```

If empty, add `GITHUB_PAT=ghp_...` to `~/.Renviron` and restart R.

### Wrong project name

Env var names are derived from `project_name` in `.datom/project.yaml`. If
your project was initialized as `"STUDY_001"`, the env vars must be
`DATOM_STUDY_001_*` — not `DATOM_Study_001_*` or `DATOM_study001_*`.
