---
title: "governance.json and the Dual-Pointer Pattern"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{governance.json and the Dual-Pointer Pattern}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

> **Companion to**: [ref.json and Always-Migration-Ready Storage](design-ref-json.html).
> Read that article first if you have not already.

The governance repo knows where the data lives -- that is the job of
`ref.json`. The reverse pointer, from the data repo back to the
governance repo, is the job of `governance.json`.

Together the two files form a **dual-pointer pattern**: each side of the
two-repo architecture carries a small JSON file that tells a reader how to
find the other side.

## Why two directions?

`ref.json` solves the reader bootstrap problem: a reader who knows the
governance repo URL and project name can always find the current data
location, even after a bucket migration.

`governance.json` solves the repair and discovery problem: given only a
data clone on disk, the tooling can locate the governance repo without
requiring the user to remember it. This enables:

- **Sync repair** -- if the git copy of `governance.json` drifts from the
  storage mirror, `.datom_sync_governance_json()` uses the git copy to push
  a fresh mirror.
- **Tooling introspection** -- scripts that walk a collection of data clones
  can discover governance without external configuration.

## Anatomy

A typical `governance.json` inside a data clone
(`.datom/governance.json`):

```json
{
  "gov_repo_url": "https://github.com/acme/datom-governance.git",
  "project_name": "STUDY_001",
  "created_at":   "2026-06-01T12:00:00Z",
  "datom_version": "0.5.0"
}
```

All four fields are always present. `gov_repo_url` and `project_name` are
the functional content; `created_at` and `datom_version` are metadata.

## Two locations, one source of truth

`governance.json` lives in two places:

| Location | Path | Purpose |
|---|---|---|
| Data git repo | `.datom/governance.json` | Canonical; version-controlled |
| Data storage | `{prefix}/datom/.metadata/governance.json` | Mirror; enables discovery without cloning |

The git copy is canonical. The storage mirror is derived and can always
be regenerated from the git copy.

```
data git repo
  .datom/
    governance.json   <-- canonical
    manifest.json
    ...

data S3 bucket (or local path)
  datom/
    .metadata/
      governance.json   <-- mirror
    tables/
      ...
```

## Lifecycle

### Created

`governance.json` is written in two situations:

1. **`datom_init_repo()`** -- when `governance` is non-NULL and the project
   has a governance store. Written in the data repo and pushed to storage
   as part of the final git + push step.

2. **`datom_attach_gov()`** -- when governance is retrofitted onto an
   existing no-gov project. Written and mirrored in the same operation.

In both cases the git copy and the storage mirror are written together and
committed in the same data-side commit.

### Repaired / re-synced

If the storage mirror is missing or stale, call:

```{r}
.datom_sync_governance_json(conn)
```

This reads the git copy from the data clone and overwrites the storage
mirror. It is idempotent and safe to call multiple times.

### Deleted

`datom_decommission()` deletes the storage mirror as part of data-storage
cleanup (step 1b). The git copy is removed with the data repo itself (step
2 -- GitHub repo deletion + local clone deletion).

## Schema

```json
{
  "gov_repo_url":  "<string> HTTPS or SSH URL of the governance GitHub repo",
  "project_name":  "<string> project name as registered in the governance repo",
  "created_at":    "<ISO 8601 UTC timestamp>",
  "datom_version": "<string> datom package version that wrote the file"
}
```

## Invariants

- The git copy is always committed before the storage mirror is pushed.
- Both copies carry identical content (same JSON bytes, same field values).
- The file is never updated after creation; it is write-once. The governance
  repo URL and project name are stable identifiers.
- The storage key is always `{prefix}/datom/.metadata/governance.json`,
  where `{prefix}` is the effective prefix from `ref.json` (post-migration
  prefix for migrated projects).

## See also

- [ref.json and Always-Migration-Ready Storage](design-ref-json.html) --
  the forward pointer from governance to data.
- [Two Repositories: Governance vs. Data](design-two-repos.html) --
  explains why the two repos are separate.
- [Credentials in Practice](credentials-in-practice.html) -- how
  `datom_store_s3_creds()` uses ref.json to spare readers from needing to
  know the data location.
