---
title: "Two Repositories: Governance vs. Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two Repositories: Governance vs. Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

> **Companion to**: [Governing a Study Portfolio](governing-a-portfolio.html).
> Read this when you wonder why STUDY-001's git repo and the gov repo
> are separate clones and not one folder inside the other.

Every datom developer has at least two git clones on their machine: a
**data clone** for the project they're currently working on, and a
**governance clone** that's shared with everyone else in the
organization. They are separate repositories, with separate histories,
separate remotes, and separate access controls.

This is a deliberate split. It's the structural choice that lets one
team govern many projects without coupling their cadences.

## What lives where

```
+-----------------------------+         +----------------------------+
|   Data repo (per project)   |         |   Governance repo (one)    |
|                             |         |                            |
|  .datom/                    |         |  projects/                 |
|    project.yaml             |         |    STUDY_001/              |
|    manifest.json            |         |      dispatch.json         |
|  table-name/                |         |      ref.json              |
|    metadata.json            |         |      migration_history.json|
|    version_history.json     |         |    STUDY_002/              |
|  README.md                  |         |      ...                   |
|                             |         |                            |
|  Audience: project team     |         |  Audience: organization    |
|  Cadence: per-write         |         |  Cadence: per-project-     |
|                             |         |  lifecycle event           |
+-----------------------------+         +----------------------------+
```

The data repo is **project-scoped**. Its commits track every
`datom_write()` and every `datom_sync()` for STUDY-001. Its readers
are STUDY-001's data engineers, plus whoever they explicitly hand the
URL to.

The governance repo is **organization-scoped**. Its commits track
projects starting and ending. Its readers are anyone who has any
business with any datom project — typically all data engineers,
analysts, and managers in the organization.

These are different lifecycles. STUDY-001 might see thirty `datom_write`
commits a month; the gov repo might see two commits in the same month
(one to register STUDY_002, one to decommission a defunct study).
Mixing them would mean every write to STUDY-001 lands in a repo that
the entire organization watches, and every governance event gets buried
in the noise of one project's daily activity.

## Why not one repo with two folders

The cleanest one-repo design would be: data project A and gov metadata
both inside `org-datom-everything.git`, in `projects/A/data/` and
`projects/A/gov/`. Tempting because there's exactly one history for
everything. We rejected it for four reasons.

**Permissions don't compose.** Project teams need write access to
their project's history. Adding a new analyst to STUDY-001 should not
also give them write access to the registry that lists every other
study in the organization. Git permissions are repo-level on every
hosting platform; splitting along the boundary makes the model
expressible.

**Pull semantics get muddy.** Today, `datom_pull()` for a project
pulls only that project's history; `datom_pull_gov()` pulls only the
registry. With one repo, every pull is everything, every time. A
manager pulling the registry to see new projects would also be
pulling every project team's recent writes — a lot of network and
disk for no benefit.

**Decommissioning leaves scars.** When STUDY-002 is decommissioned,
its data history goes with it (the data git repo gets deleted along
with the bucket contents). The gov repo records the
decommissioning event but doesn't have to carry the corpse of every
deleted project's commits. With one repo, every decommissioned project
either lives forever in history or gets surgically excised — both bad.

**The companion-package boundary becomes ambiguous.** datom's
governance code is on its way out of this package. A future companion
package (working name: `daapr`) will own gov-side operations —
registry queries, cross-project audit, governance store administration
— while datom keeps the per-project data operations. The boundary is
expressible *because* the artifacts are already split. Functions in
[R/utils-gov.R](https://github.com/amashadihossein/datom) tagged
`# GOV_SEAM:` mark exactly the surface that will move.

## How the split shows up in API

The split is visible to you whenever you set up a store:

```{r}
store <- datom_store(
  governance     = gov_component,    # 1. governance backend
  data           = data_component,   # 2. data backend
  github_pat     = pat,
  data_repo_url  = "https://...",    # 3. data git remote
  gov_repo_url   = "https://...",    # 4. gov git remote
  gov_local_path = "/path/to/clone"  # 5. gov local clone
)
```

Five fields, two for each repo plus one for the shared GitHub PAT.
The repetition is the cost of the split. The benefits — lifecycle
independence, permission boundaries, decommission safety, future
companion-package handoff — are why the cost is worth paying.

For readers, the cost is lower: they don't see a data clone at all
(they read parquet bytes directly from the data store) and the gov
clone is implicit (datom downloads what it needs over the network).
A reader's `datom_store(...)` call has fewer fields because most of
the boilerplate is for engineering, not consumption.

## The governance store is two things

A subtlety worth surfacing: the **governance store** is *both* a git
repository (the registry, the source of truth) *and* an object-store
location (where dispatch/ref/migration JSONs are also written). They
are kept synchronized by datom. The git copy is the canonical version;
the object-store copy is what readers without a clone reach for.

This is symmetrical to the data side: data git repo (metadata, code,
versioned commits) plus data object store (the actual parquet bytes).
In both cases, git is canonical for things humans should review;
object storage is canonical for things machines should fetch quickly.

The fact that the two halves of "the gov repo" and the two halves of
"the data repo" follow the same shape is not coincidence — it's the
core datom pattern applied at two scopes. See [The datom Model: Code
in Git, Data in Cloud](design-datom-model.html) for the underlying
principle.

## What a future companion package looks like

If you read [R/utils-gov.R](https://github.com/amashadihossein/datom)
and look for the `# GOV_SEAM:` markers, you're looking at the future
boundary. Every function tagged that way is a candidate to move out
of datom and into the companion package. The contract is simple:

- All gov-write code lives in functions tagged `# GOV_SEAM:`.
- No data-side code path calls a `# GOV_SEAM:` function directly —
  the only callers are public datom functions whose own purpose is
  governance-adjacent (`datom_init_repo`, `datom_decommission`,
  `datom_init_gov`).
- The companion package, when it ships, will reimplement these
  functions and depend on datom for the data-side operations only.

The split into two git repos was the prerequisite. Without it, "lift
the governance code into a companion package" would mean rewriting
how every commit is structured. With it, the move is closer to
mechanical.

## Where this leads

The two-repo split is the structural counterpart to two design
choices already established:

- The metadata-vs-data split — see [The datom Model: Code in Git,
  Data in Cloud](design-datom-model.html).
- The serverless property that follows from "everything is git +
  object storage" — see [Serverless and Distributed by
  Design](design-serverless.html).

All three are different views of the same underlying choice: keep
the moving parts small, keep their interfaces explicit, and let the
existing tools (git, S3) do the heavy lifting.
