---
title: "First Extract"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First Extract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{=html}
<link rel="stylesheet" href="../inst/vignette-setup/datom.css">
```

<div class="datom-callout datom-callout-goal" markdown="1">
**Goal:** Stand up a versioned datom project using a local filesystem store --
write your first
table, read it back, and confirm git is tracking the history. No AWS account
needed. The same workflow extends to a shared team space (S3 or a shared
filesystem) without changing a single function call.
</div>

> **Already done the two-minute tour in the README?** The tour and this
> article cover the same ground. If your project initialized cleanly and
> `datom_read()` returned `TRUE`, you can jump straight to
> [Month 2 Arrives](month-2-arrives.html).

You are the data engineer for **STUDY-001**, a Phase II clinical trial. The first
EDC extract has just landed. datom lets you and your collaborators build a
**shared, versioned data space** -- multiple engineers writing new extracts,
multiple analysts reading any version, all coordinated through a single git
history. Every write is a git commit; every read resolves to an exact version
SHA. No one can silently overwrite history, and anyone with access to the repo
can reproduce any past analysis by pinning to a SHA.

This first article walks the local-only path. The same workflow -- same
functions, same commands -- works for a shared S3 space: you build an S3 store
instead of a local one (see [Starting on S3](start-on-s3.html)).

## Requirements

datom keeps **metadata in git** (diff-able, auditable) and **data wherever you
tell it to live** (S3 or a local directory). Even when data lives on a local
filesystem, metadata still goes to a git remote -- that is how version history stays
reproducible across machines.

You need two things, both one-time:

- **A GitHub account** with a personal access token (PAT) scoped to `repo`.
  Store it in your OS keychain once with `keyring::key_set("GITHUB_PAT")`;
  every article after this picks it up automatically. Using `keyring` keeps the
  PAT out of your code and command history.
- **The `gh` CLI is not required** -- datom creates GitHub repos through the
  GitHub REST API directly using your PAT.

No AWS, no cloud account, no governance repo for this article.

<div class="datom-callout datom-callout-tip" markdown="1">
**Verify your keyring setup** before continuing:

```r
nzchar(keyring::key_get("GITHUB_PAT"))   # should return TRUE
```

If it errors, set the PAT with `keyring::key_set("GITHUB_PAT")` first.
</div>

## Configure your environment

Everything machine-specific lives in one block. Set these values once for your
environment; every code chunk after this is copy/paste as-is.

```{r}
library(datom)
library(fs)

# --- Settings you control --------------------------------------------------
project_name <- "STUDY_001"          # logical project name (recorded in metadata)
repo_name    <- "study-001-data"     # GitHub repo name for the metadata repo

# dev_dir  -- your local clone of the metadata git repo (stays on your machine)
# data_dir -- where parquet bytes are written; point at a shared location
#             (network mount, or an S3 store) for a real team space. Temp dirs
#             are used here so the article leaves nothing behind.
dev_dir  <- path(tempdir(), "study_001_dev")
data_dir <- path(tempdir(), "study_001_data")

# Your GitHub personal access token (PAT), scoped to `repo`. Stored once in your
# OS keychain with keyring::key_set("GITHUB_PAT"); read here by name so the
# token never appears in your code or command history.
github_pat <- keyring::key_get("GITHUB_PAT")
# ---------------------------------------------------------------------------

dir_create(data_dir)
```

## Build a store

A **store** bundles the addresses datom needs: where parquet bytes go and the
GitHub PAT that lets datom push metadata. Governance is not attached
(`governance = NULL`); it is opt-in and added later through the governance
companion package when sharing across a portfolio matters.

```{r}
store <- datom_store(
  governance = NULL,
  data       = datom_store_local(path = data_dir),
  github_pat = github_pat
)
```

## Initialize the data repository

```{r}
datom_init_repo(
  path         = dev_dir,
  project_name = project_name,
  store        = store,
  create_repo  = TRUE,
  repo_name    = repo_name
)
```

This creates a GitHub repo, clones it into `dev_dir`, and commits a
`project.yaml` that records the project's data store address. The git repo is
now live on GitHub. No parquet data is pushed to GitHub -- only the metadata
commits travel over the wire; the parquet bytes stay in `data_dir`.

Take a moment to inspect the repo structure before moving on:

```{r}
# Metadata layout in the git clone
fs::dir_tree(dev_dir)

# Storage layout (empty until first write)
fs::dir_tree(data_dir)
```

## Connect

```{r}
conn <- datom_get_conn(path = dev_dir, store = store)
print(conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "developer"
#> * Backend: "local"
#> * Root: "/tmp/.../study_001_data"
#> * Path: "/tmp/.../study_001_dev"
#> * Governance: not attached
```

## Write your first extract

The month-1 extract has just landed. Load the demographics snapshot for
subjects enrolled by 2026-01-28:

```{r}
dm_m1 <- datom_example_data("dm", cutoff_date = "2026-01-28")
nrow(dm_m1)
#> [1] 4
```

Write it as a versioned datom table:

```{r}
datom_write(
  conn,
  data    = dm_m1,
  name    = "dm",
  message = "Initial DM extract through 2026-01-28"
)
#> v Wrote "dm" (full): "a8ee7a31"
```

Three things just happened, in this order:

1. The data frame was serialized to parquet and written to `data_dir`.
   **No data was pushed to the GitHub repo** -- parquet bytes never leave
   your local store.
2. `metadata.json` and `version_history.json` were updated in the git clone
   and committed.
3. The metadata commit was pushed to GitHub. The version is now auditable from
   any machine with repo access, but the raw data stays where you put it.

After the write, explore what changed:

```{r}
fs::dir_tree(data_dir)    # parquet file now present
datom_status(conn)        # table list with SHAs
datom_history(conn, "dm") # version history
```

## Read it back as a reader

Writing is a developer action. Reading is what analysts and downstream
pipelines do, and they use the **reader role** -- a connection built from the
data store alone, with no GitHub PAT and no local clone. You build a reader
store by leaving `github_pat` unset, then connect by project name:

```{r}
reader_store <- datom_store(
  governance = NULL,
  data       = datom_store_local(path = data_dir)   # same data location
)                                                    # no PAT -> reader role

reader_conn <- datom_get_conn(store = reader_store, project_name = project_name)
print(reader_conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "reader"
#> * Backend: "local"
```

Read the table back through the reader connection:

```{r}
dm_back <- datom_read(reader_conn, "dm")
```

datom stores data in [Apache Parquet](https://parquet.apache.org/) format and
reads it back as a `tibble`. Compare as a tibble for a clean round-trip check:

```{r}
identical(datom_read(reader_conn, "dm"), tibble::as_tibble(dm_m1))
#> [1] TRUE
```

The read does **not** go through GitHub. The reader resolves the parquet file
from the data store and streams it back -- exactly the path a teammate on
another machine takes. They need access to the data store, not to the git repo.

## Where you are

You have a fully versioned datom project up and running:

- One table (`dm`) with one version, one parquet file in `data_dir`
- Metadata committed and pushed to GitHub -- version history is auditable
- Parquet data stays local -- **nothing sensitive went to GitHub**
- No governance attached; it is opt-in and added later through the governance
  companion package when sharing across a portfolio matters

In the next article, the **month-2 extract arrives** with new subjects, and you
write a second version without overwriting the first.

## Carry on, or tear down

**Continuing to [Month 2 Arrives](month-2-arrives.html)?** Stop here. Keep your
R session open and reuse `conn`, `store`, and your settings (`dev_dir`,
`data_dir`, `project_name`) in the next article -- do **not** tear anything
down. If you closed your session, article 2 opens with the same configuration
block so you can reconnect.

**Done exploring?** Tear the project down. Pick the one option that fits, and
copy only that chunk:

```{r}
# Option A -- full scripted teardown (deletes local files AND the GitHub repo).
# Do this BEFORE any manual unlink().
datom_repo_delete(conn, confirm = "STUDY_001")
```

```{r}
# Option B -- local only (the GitHub repo stays; delete it from the UI later).
unlink(c(dev_dir, data_dir), recursive = TRUE)
```

Do not call `unlink()` before `datom_repo_delete()` -- removing the local
clone first strips the GitHub remote reference and the remote repo will not
be deleted.
