---
title: "First Extract"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First Extract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{=html}
<link rel="stylesheet" href="../inst/vignette-setup/datom.css">
```

<div class="datom-callout datom-callout-goal" markdown="1">
**Goal:** Stand up a versioned datom project on your laptop — write your first
table, read it back, and confirm git is tracking the history. No AWS account
needed. The same workflow extends to a shared team space (S3 or a shared
filesystem) without changing a single function call.
</div>

You are the data engineer for **STUDY-001**, a Phase II clinical trial. The first
EDC extract has just landed. datom lets you and your collaborators build a
**shared, versioned data space** — multiple engineers writing new extracts,
multiple analysts reading any version, all coordinated through a single git
history. Every write is a git commit; every read resolves to an exact version
SHA. No one can silently overwrite history, and anyone with access to the repo
can reproduce any past analysis by pinning to a SHA.

This first article walks the local-only path. The same workflow — same
functions, same commands — works for a shared S3 space once you swap the local
store for an S3 store in [Promoting to S3](promoting-to-s3.html).

## Requirements

datom keeps **metadata in git** (diff-able, auditable) and **data wherever you
tell it to live** (S3 or a local directory). Even when data lives on your
laptop, metadata still goes to a git remote — that is how version history stays
reproducible across machines.

You need two things, both one-time:

- **A GitHub account** with a personal access token (PAT) scoped to `repo`.
  Store it in your OS keychain once with `keyring::key_set("GITHUB_PAT")`;
  every article after this picks it up automatically.
  See [Credentials in Practice](credentials-in-practice.html) for a step-by-step
  walkthrough of PAT creation and keyring setup.
- **The `gh` CLI is not required** — datom creates GitHub repos through the
  GitHub REST API directly using your PAT.

No AWS, no cloud account, no governance repo for this article.

<div class="datom-callout datom-callout-tip" markdown="1">
**Verify your keyring setup** before continuing:

```r
nzchar(keyring::key_get("GITHUB_PAT"))   # should return TRUE
```

If it errors, follow the [Credentials in Practice](credentials-in-practice.html)
setup steps first.
</div>

## Set up your working paths

Two paths are needed, and they serve different roles:

- **`study_dir`** — the local clone of the data git repository. This is
  where metadata (`project.yaml`, `metadata.json`, version history) lives.
  In a team setting this would be cloned on every developer's machine.
- **`data_root`** — the directory where parquet bytes are written. In a team
  setting this would be an S3 bucket (or a shared network mount), so every
  team member reads from the same physical store.

Here both point to temporary directories for demonstration. Replace them with
real paths — or an S3 store — when you are ready for a persistent shared space.

```{r}
library(datom)
library(fs)

study_dir <- path(tempdir(), "study_001_data")        # data git clone
data_root <- path(tempdir(), "study_001_data_root")   # parquet bytes live here

dir_create(data_root)
```

## Build a store

A **store** bundles the addresses datom needs: where parquet bytes go and the
GitHub PAT that lets datom push metadata. Governance is not attached yet
(`governance = NULL`); you'll add it in [Promoting to S3](promoting-to-s3.html).

```{r}
data_component <- datom_store_local(path = data_root)

store <- datom_store(
  governance = NULL,
  data       = data_component,
  github_pat = keyring::key_get("GITHUB_PAT")
)
```

## Initialize the data repository

```{r}
datom_init_repo(
  path         = study_dir,
  project_name = "STUDY_001",
  store        = store,
  create_repo  = TRUE,
  repo_name    = "study-001-data"
)
```

This creates a GitHub repo, clones it into `study_dir`, and commits a
`project.yaml` that records the project's data store address. The git repo is
now live on GitHub. No parquet data is pushed to GitHub — only the metadata
commits travel over the wire; the parquet bytes stay in `data_root`.

Take a moment to inspect the repo structure before moving on:

```{r}
# Metadata layout in the git clone
fs::dir_tree(study_dir)

# Storage layout (empty until first write)
fs::dir_tree(data_root)
```

## Connect

```{r}
conn <- datom_get_conn(path = study_dir, store = store)
print(conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "developer"
#> * Backend: "local"
#> * Root: "/tmp/.../study_001_data_root"
#> * Path: "/tmp/.../study_001_data"
#> * Governance: not attached
```

## Write your first extract

The month-1 extract has just landed. Load the demographics snapshot for
subjects enrolled by 2026-01-28:

```{r}
dm_m1 <- datom_example_data("dm", cutoff_date = "2026-01-28")
nrow(dm_m1)
#> [1] 6
```

Write it as a versioned datom table:

```{r}
datom_write(
  conn,
  data    = dm_m1,
  name    = "dm",
  message = "Initial DM extract through 2026-01-28"
)
#> v Wrote "dm" (full): "a8ee7a31"
```

Three things just happened, in this order:

1. The data frame was serialized to parquet and written to `data_root`.
   **No data was pushed to the GitHub repo** — parquet bytes never leave
   your local store.
2. `metadata.json` and `version_history.json` were updated in the git clone
   and committed.
3. The metadata commit was pushed to GitHub. The version is now auditable from
   any machine with repo access, but the raw data stays where you put it.

After the write, explore what changed:

```{r}
fs::dir_tree(data_root)   # parquet file now present
datom_status(conn)        # table list with SHAs
datom_history(conn, "dm") # version history
```

## Read it back

```{r}
dm_back <- datom_read(conn, "dm")
```

datom stores data in [Apache Parquet](https://parquet.apache.org/) format and
reads it back as a `tibble`. If your original object was not already a tibble,
the classes will differ even though the data is identical:

```{r}
identical(datom_read(conn, "dm"), dm_m1)
#> [1] FALSE  # dm_m1 may carry extra classes (e.g. data.frame)

identical(datom_read(conn, "dm"), tibble::as_tibble(dm_m1))
#> [1] TRUE   # compare as tibble for a clean round-trip check
```

The read does **not** go through GitHub. It uses the manifest cached in
`data_root` to locate the parquet file and stream it back. This is the same
path a data reader on a different machine takes — they need access to the data
store, not to the git repo.

## Where you are

You have a fully versioned datom project on your laptop:

- One table (`dm`) with one version, one parquet file in `data_root`
- Metadata committed and pushed to GitHub — version history is auditable
- Parquet data stays local — **nothing sensitive went to GitHub**
- No governance attached yet; you'll add it in
  [Promoting to S3](promoting-to-s3.html) when sharing matters

In the next article, the **month-2 extract arrives** with new subjects, and you
write a second version without overwriting the first.

## Teardown

When you are done exploring, remove all local artefacts:

```{r}
unlink(c(study_dir, data_root), recursive = TRUE)
```

This removes only your local files. The GitHub repo `study-001-data` remains on
GitHub — delete it from the GitHub UI when you no longer need it. Article 7
introduces `datom_decommission()`, which handles the full scripted teardown
including the remote repo.
