---
title: "First Extract"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First Extract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{=html}
<link rel="stylesheet" href="../inst/vignette-setup/datom.css">
```

<div class="datom-callout datom-callout-goal" markdown="1">
**Goal:** Stand up a versioned datom project using a local filesystem store and
onboard data using `datom_sync()` -- the primary workflow for bringing files
into datom. By the end of this article the project holds multiple versioned
tables, and you have seen: syncing one file, updating it, the idempotent no-op,
and syncing a batch. No AWS account needed. The same workflow extends unchanged
to S3 (see [Starting on S3](start-on-s3.html)).
</div>

> **Already done the two-minute tour in the README?** The tour and this
> article cover the same ground. If your project initialized cleanly and
> `datom_read()` returned `TRUE`, you can jump straight to
> [Starting on S3](start-on-s3.html) or
> [Tracing Data Lineage](source-lineage.html).

You are the data engineer for **STUDY-001**, a Phase II clinical trial. EDC
extracts land as files in a staging folder; your job is to onboard them into a
**shared, versioned data space** -- multiple engineers writing new extracts,
multiple analysts reading any version, all coordinated through a single git
history. Every sync is a git commit; every read resolves to an exact version
SHA. No one can silently overwrite history, and anyone with access to the repo
can reproduce any past analysis by pinning to a SHA.

This first article walks the local-only path. The same workflow -- same
functions, same commands -- works for a shared S3 space: you build an S3 store
instead of a local one (see [Starting on S3](start-on-s3.html)).

<div class="datom-callout datom-callout-tip" markdown="1">
**Two locations, two roles:**

- **Input folder** (mutable, disposable) -- a landing zone for data not yet
  onboarded. Files here can be overwritten or deleted freely. Re-syncing
  already-onboarded content is a no-op. Nothing here is the source of truth.
- **Storage** (immutable, permanent) -- once `datom_sync()` onboards a file,
  storage holds the versioned, content-addressed copy. The original input file
  is no longer needed; the onboarded data stands on its own.
</div>

## Requirements

datom keeps **metadata in git** (diff-able, auditable) and **data wherever you
tell it to live** (S3 or a local directory). Even when data lives on a local
filesystem, metadata still goes to a git remote -- that is how version history
stays reproducible across machines.

You need two things, both one-time:

- **A GitHub account** with a personal access token (PAT) scoped to `repo`.
  Store it in your OS keychain once with `keyring::key_set("GITHUB_PAT")`;
  every article after this picks it up automatically. Using `keyring` keeps the
  PAT out of your code and command history.
- **The `gh` CLI is not required** -- datom creates GitHub repos through the
  GitHub REST API directly using your PAT.

No AWS, no cloud account, no governance repo for this article.

<div class="datom-callout datom-callout-tip" markdown="1">
**Verify your keyring setup** before continuing:

```r
nzchar(keyring::key_get("GITHUB_PAT"))   # should return TRUE
```

If it errors, set the PAT with `keyring::key_set("GITHUB_PAT")` first.
</div>

## Configure your environment

Everything machine-specific lives in one block. Set these values once for your
environment; every code chunk after this is copy/paste as-is.

```{r}
library(datom)
library(fs)

# --- Settings you control --------------------------------------------------
project_name <- "STUDY_001"          # logical project name (recorded in metadata)
repo_name    <- "study-001-data"     # GitHub repo name for the metadata repo

# dev_dir  -- your local clone of the metadata git repo (stays on your machine)
# data_dir -- where parquet bytes are written; point at a shared location
#             (network mount, or an S3 store) for a real team space. Temp dirs
#             are used here so the article leaves nothing behind.
dev_dir  <- path(tempdir(), "study_001_dev")
data_dir <- path(tempdir(), "study_001_data")

# Your GitHub personal access token (PAT), scoped to `repo`. Stored once in your
# OS keychain with keyring::key_set("GITHUB_PAT"); read here by name so the
# token never appears in your code or command history.
github_pat <- keyring::key_get("GITHUB_PAT")
# ---------------------------------------------------------------------------

dir_create(data_dir)
```

## Build a store

A **store** bundles the addresses datom needs: where parquet bytes go and the
GitHub PAT that lets datom push metadata. Governance is not attached
(`governance = NULL`); it is opt-in and added later through the governance
companion package when sharing across a portfolio matters.

```{r}
store <- datom_store(
  governance = NULL,
  data       = datom_store_local(path = data_dir),
  github_pat = github_pat
)
```

## Initialize the data repository

```{r}
datom_init_repo(
  path         = dev_dir,
  project_name = project_name,
  store        = store,
  create_repo  = TRUE,
  repo_name    = repo_name
)
```

This creates a GitHub repo, clones it into `dev_dir`, and commits a
`project.yaml` that records the project's data store address. The git repo is
now live on GitHub. No parquet data is pushed to GitHub -- only the metadata
commits travel over the wire; the parquet bytes stay in `data_dir`.

`datom_init_repo()` also creates an `input_files/` directory inside the clone.
It is gitignored -- files placed there are never committed. It is the inbox for
`datom_sync()`.

Take a moment to inspect the repo structure before moving on:

```{r}
# Metadata layout in the git clone (note input_files/ -- your sync inbox)
fs::dir_tree(dev_dir)

# Storage layout (empty until first sync)
fs::dir_tree(data_dir)
```

## Connect

```{r}
conn <- datom_get_conn(path = dev_dir, store = store)
print(conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "developer"
#> * Backend: "local"
#> * Root: "/tmp/.../study_001_data"
#> * Path: "/tmp/.../study_001_dev"
#> * Governance: not attached
```

## Step 1: Sync one file

The month-1 extract has just landed -- a single demographics CSV. Drop it
in the input folder:

```{r}
# The input folder lives inside the git clone but is gitignored.
# Files placed here are the raw material for datom_sync().
input_dir <- path(dev_dir, "input_files")

write.csv(
  datom_example_data("dm", cutoff_date = "2026-01-28"),
  path(input_dir, "dm.csv"),
  row.names = FALSE
)
```

Scan the folder to build a sync manifest -- datom detects what is new, changed,
or unchanged relative to what has already been onboarded:

```{r}
manifest <- datom_sync_manifest(conn)
#> i Scanned 1 file: 1 new, 0 changed, 0 unchanged.
manifest
#>   name  file         format file_sha status
#> 1   dm  .../dm.csv   csv    7a3b...  new
```

Now onboard it:

```{r}
datom_sync(conn, manifest)
#> i Syncing 1 table...
#> v dm synced (new).
#> i Sync complete: 1 succeeded, 0 failed, 0 skipped.
```

Three things just happened, in this order:

1. The CSV was read and serialized to parquet in `data_dir`.
   **No data was pushed to the GitHub repo** -- parquet bytes never leave
   your local store.
2. `metadata.json` and `version_history.json` were updated in the git clone
   and committed.
3. The metadata commit was pushed to GitHub. The version is now auditable from
   any machine with repo access, but the raw data stays where you put it.

Confirm with `datom_list()`, `datom_history()`, and a round-trip read:

```{r}
datom_list(conn)
#>   name current_version current_data_sha last_updated
#> 1   dm        a8ee7a31         4b6d0a7e 2026-01-28T...

datom_history(conn, "dm")
#>    version  data_sha timestamp            message
#> 1 a8ee7a31 4b6d0a7e 2026-01-28T09:02:11Z dm synced from dm.csv

datom_read(conn, "dm")
```

**Key point:** the input file is now disposable. Delete it and the data is
still accessible from storage -- storage is the source of truth, not the input
folder:

```{r}
file_delete(path(input_dir, "dm.csv"))
datom_read(conn, "dm")   # still works -- storage is the permanent record
```

## Step 2: Update one file

A month passes. The month-2 extract arrives with new subjects. Overwrite the
input file and sync again:

```{r}
write.csv(
  datom_example_data("dm", cutoff_date = "2026-02-28"),
  path(input_dir, "dm.csv"),
  row.names = FALSE
)

manifest <- datom_sync_manifest(conn)
#> i Scanned 1 file: 0 new, 1 changed, 0 unchanged.

datom_sync(conn, manifest)
#> i Syncing 1 table...
#> v dm synced (changed).
#> i Sync complete: 1 succeeded, 0 failed, 0 skipped.
```

The table now has two versions. The previous version is still intact -- nothing
was overwritten:

```{r}
datom_history(conn, "dm")
#>    version  data_sha timestamp            message
#> 1 5c1a3f7b 9e8f1c2d 2026-02-28T10:14:02Z dm synced from dm.csv
#> 2 a8ee7a31 4b6d0a7e 2026-01-28T09:02:11Z dm synced from dm.csv

# Read the current version (month 2)
nrow(datom_read(conn, "dm"))
#> [1] 16

# Read the prior version (month 1) by its SHA
hist   <- datom_history(conn, "dm")
m1_ver <- hist$version[nrow(hist)]   # oldest row is the month-1 version
nrow(datom_read(conn, "dm", version = m1_ver))
#> [1] 4
```

Both versions coexist; any historical snapshot is retrievable by its SHA.
This is the property that downstream statisticians, regulators, and auditors
rely on.

## Step 3: No-op repeat

Run `datom_sync()` again with nothing changed:

```{r}
manifest <- datom_sync_manifest(conn)
#> i Scanned 1 file: 0 new, 0 changed, 1 unchanged.

datom_sync(conn, manifest)
#> i All files unchanged. Nothing to sync.
```

Re-syncing identical content is a no-op. This is what makes sync safe to run
in a scheduled job or pipeline -- accidental re-runs cost nothing and pollute
no history. The same idempotency applies to `datom_write()`: writing an
identical data frame to the same table name detects the duplicate and skips.

## Step 4: A batch of files

It is now month 3. The data management team has switched from emailing single
files to dropping a folder of monthly extracts. Today's drop contains four
domains:

```{r}
cutoff <- "2026-03-28"

write.csv(datom_example_data("dm", cutoff_date = cutoff),
          path(input_dir, "dm.csv"), row.names = FALSE)
write.csv(datom_example_data("ex", cutoff_date = cutoff),
          path(input_dir, "ex.csv"), row.names = FALSE)
write.csv(datom_example_data("lb", cutoff_date = cutoff),
          path(input_dir, "lb.csv"), row.names = FALSE)
write.csv(datom_example_data("ae", cutoff_date = cutoff),
          path(input_dir, "ae.csv"), row.names = FALSE)
```

Scan and sync in one pass:

```{r}
manifest <- datom_sync_manifest(conn)
#> i Scanned 4 files: 3 new, 1 changed, 0 unchanged.
manifest
#>   name  file         format file_sha status
#> 1   dm  .../dm.csv   csv    c41a...  changed
#> 2   ex  .../ex.csv   csv    9b08...  new
#> 3   lb  .../lb.csv   csv    72d3...  new
#> 4   ae  .../ae.csv   csv    1e4a...  new

datom_sync(conn, manifest)
#> i Syncing 4 tables...
#> v dm synced (changed).
#> v ex synced (new).
#> v lb synced (new).
#> v ae synced (new).
#> i Sync complete: 4 succeeded, 0 failed, 0 skipped.
```

`dm` is `changed` (month-3 update); `ex`, `lb`, `ae` are brand new. All four
are now versioned:

```{r}
datom_list(conn)
#>   name current_version current_data_sha last_updated
#> 1   ae        3a17b8e2         e91d04ff 2026-03-28T...
#> 2   dm        d0922fc7         c2e80a14 2026-03-28T...
#> 3   ex        f44910b5         88a73e02 2026-03-28T...
#> 4   lb        718e02ca         4c3812dd 2026-03-28T...
```

## Project health checks

`datom_status()` summarizes the project state at a glance:

```{r}
datom_status(conn)
#> -- datom status: STUDY_001
#> v Git: clean, in sync with origin
#> i Tables on local: 4
#> i Last commit: <sha> "Update ae"
```

`datom_validate()` cross-checks that every table in the manifest has its
parquet file in the data store and its metadata in git history:

```{r}
datom_validate(conn)
#> v 4 tables validated.
#> v Manifest <-> data store: consistent.
#> v Manifest <-> git history: consistent.
```

If you ever see a discrepancy from `datom_validate()`, that is the moment to
stop and investigate before doing more work -- it means the project's state
has drifted from one of git or storage.

## Reading as a reader

Writing is a developer action. Reading is what analysts and downstream
pipelines do, and they use the **reader role** -- a connection built from the
data store alone, with no GitHub PAT and no local clone. You build a reader
store by leaving `github_pat` unset, then connect by project name:

```{r}
reader_store <- datom_store(
  governance = NULL,
  data       = datom_store_local(path = data_dir)   # same data location
)                                                    # no PAT -> reader role

reader_conn <- datom_get_conn(store = reader_store, project_name = project_name)
print(reader_conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "reader"
#> * Backend: "local"
```

Read any table through the reader connection:

```{r}
datom_read(reader_conn, "lb")   # labs, current version from storage
```

datom stores data in [Apache Parquet](https://parquet.apache.org/) format and
reads it back as a `tibble`. The read does **not** go through GitHub -- the
reader resolves the parquet file directly from the data store. A teammate on
another machine takes the same path: they need access to the data store, not
to the git repo.

## Writing a computed table with `datom_write()`

`datom_sync()` is the primary workflow for onboarding files from a staging
folder. For tables derived in code -- where the data frame is computed rather
than imported from a file -- use `datom_write()` directly:

```{r}
# Derive a summary table in code
lb <- datom_read(conn, "lb")
lb_summary <- dplyr::summarise(
  dplyr::group_by(lb, LBTESTCD),
  n = dplyr::n(),
  .groups = "drop"
)

# Write it as a versioned table, declaring its parent for lineage
datom_write(
  conn,
  data    = lb_summary,
  name    = "lb_summary",
  message = "Lab test counts from month-3 LB",
  parents = list(
    list(table = "lb", version = datom_history(conn, "lb")$version[1])
  )
)
#> v Wrote "lb_summary" (full): "b9c4e21a"
```

`datom_write()` takes the same connection and produces the same versioned,
content-addressed result. The `parents` argument declares lineage -- this table
was derived from `lb` -- making the provenance chain auditable (see
[Tracing Data Lineage](source-lineage.html)).

## Where you are

- A full sync-based onboarding workflow: one file, update, no-op, batch.
- Four raw tables versioned; one derived table written directly.
- Input files are disposable; storage is the permanent record.
- `datom_status()` and `datom_validate()` provide consistency checks.
- Readers access data from storage alone -- no git, no PAT.

From here, [Starting on S3](start-on-s3.html) shows the same workflow with data
in object storage, and [Tracing Data Lineage](source-lineage.html) shows how
derived tables record where they came from.

## Teardown

When done, remove the project. Pick the one option that fits, and copy only
that chunk:

```{r}
# Option A -- full scripted teardown (deletes local files AND the GitHub repo).
# Do this BEFORE any manual unlink().
datom_repo_delete(conn, confirm = "STUDY_001")
```

```{r}
# Option B -- local only (the GitHub repo stays; delete it from the UI later).
unlink(c(dev_dir, data_dir), recursive = TRUE)
```

Do not call `unlink()` before `datom_repo_delete()` -- removing the local
clone first strips the GitHub remote reference and the remote repo will not
be deleted.
