---
title: "A Folder of Extracts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Folder of Extracts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

> **Where we left off:** STUDY-001 has one table (`dm`) with two versions --
> month-1 and month-2 extracts.

It is now mid-March. The data management team has switched from emailing single
CSVs to dropping a **folder** of monthly extracts. Today's drop contains four
domains: `dm`, `ex`, `lb` (labs), and `ae` (adverse events).

This article introduces three capabilities: scanning an input folder
(`datom_sync_manifest()`), syncing all changed files at once
(`datom_sync()`), and asking datom for the consistency state of the
project (`datom_status()`, `datom_validate()`).

## Reconnect (only if you closed your session)

If you've been working in the same R session since article 1 or 2, you already
have `conn` and `store` -- skip ahead to the drop below.

If you closed your session, re-run the configuration block to reconnect:

```{r}
library(datom)
library(fs)

# The same machine-specific settings from the earlier articles.
project_name <- "STUDY_001"
dev_dir      <- path(tempdir(), "study_001_dev")
data_dir     <- path(tempdir(), "study_001_data")
github_pat   <- keyring::key_get("GITHUB_PAT")

store <- datom_store(
  governance = NULL,
  data       = datom_store_local(path = data_dir),
  github_pat = github_pat
)

conn <- datom_get_conn(path = dev_dir, store = store)
```

## Drop the extracts in `input_files/`

`datom_init_repo()` created an `input_files/` directory inside your data
clone. It is gitignored -- files placed there are never committed; they're
the inbox for `datom_sync()`.

```{r}
library(datom)
library(fs)

input_dir <- path(dev_dir, "input_files")
cutoff    <- "2026-03-28"

write.csv(datom_example_data("dm", cutoff_date = cutoff),
          path(input_dir, "dm.csv"), row.names = FALSE)
write.csv(datom_example_data("ex", cutoff_date = cutoff),
          path(input_dir, "ex.csv"), row.names = FALSE)
write.csv(datom_example_data("lb", cutoff_date = cutoff),
          path(input_dir, "lb.csv"), row.names = FALSE)
write.csv(datom_example_data("ae", cutoff_date = cutoff),
          path(input_dir, "ae.csv"), row.names = FALSE)
```

## Scan the folder

```{r}
manifest <- datom_sync_manifest(conn)
#> i Scanned 4 files: 3 new, 1 changed, 0 unchanged.
manifest
#>   name       file format file_sha status
#> 1   dm  .../dm.csv    csv  c41a... changed
#> 2   ex  .../ex.csv    csv  9b08... new
#> 3   lb  .../lb.csv    csv  72d3... new
#> 4   ae  .../ae.csv    csv  1e4a... new
```

The manifest classifies each file:

- **new** -- datom has never seen this table.
- **changed** -- the table exists, but the source file has a new SHA.
- **unchanged** -- same source file as last time. Sync will skip these.

`dm` is `changed` because article 2 wrote month-2; this is the month-3
update. `ex`, `lb`, and `ae` are brand-new tables.

## Sync everything

```{r}
results <- datom_sync(conn, manifest)
#> i Syncing 4 tables...
#> v dm synced (changed).
#> v ex synced (new).
#> v lb synced (new).
#> v ae synced (new).
#> i Sync complete: 4 succeeded, 0 failed, 0 skipped.

results
#>   name       file format ... status   result error
#> 1   dm  .../dm.csv    csv     changed success  <NA>
#> 2   ex  .../ex.csv    csv     new     success  <NA>
#> 3   lb  .../lb.csv    csv     new     success  <NA>
#> 4   ae  .../ae.csv    csv     new     success  <NA>
```

Each `datom_sync()` row becomes one git commit and one parquet upload.
`datom_history()` on any of the four tables now shows its first or
incremented version.

## What's in the project?

```{r}
datom_list(conn)
#>   name current_version current_data_sha last_updated
#> 1   ae         3a17b8e2         e91d04ff 2026-03-28T...
#> 2   dm         d0922fc7         c2e80a14 2026-03-28T...
#> 3   ex         f44910b5         88a73e02 2026-03-28T...
#> 4   lb         718e02ca         4c3812dd 2026-03-28T...
```

## Check the project's status

`datom_status()` summarizes what's in your local clone vs. what's pushed:

```{r}
datom_status(conn)
#> -- datom status: STUDY_001
#> v Git: clean, in sync with origin
#> i Tables on local: 4
#> i Last commit: <sha> "Update ae"
```

`datom_validate()` goes a step further -- it cross-checks that every table in
the local manifest has its parquet file in the data store and its metadata
in git history:

```{r}
datom_validate(conn)
#> v 4 tables validated.
#> v Manifest <-> data store: consistent.
#> v Manifest <-> git history: consistent.
```

If you ever see a discrepancy from `datom_validate()`, that's the moment to
stop and investigate before doing more work -- it means the project's state
has drifted from one of git or storage.

## Re-running is safe

If you run `datom_sync_manifest()` immediately after a successful sync,
every file will be `unchanged` and `datom_sync()` will report nothing to do:

```{r}
datom_sync_manifest(conn)
#> i Scanned 4 files: 0 new, 0 changed, 4 unchanged.

datom_sync(conn, datom_sync_manifest(conn))
#> i All files unchanged. Nothing to sync.
```

This is the property that lets sync run in a cron job without surprises.

## Read as a reader

Any of the four tables is now readable through a **reader** connection -- no
PAT, no local clone, just the data store:

```{r}
reader_store <- datom_store(
  governance = NULL,
  data       = datom_store_local(path = data_dir)
)
reader_conn <- datom_get_conn(store = reader_store, project_name = project_name)

datom_read(reader_conn, "lb")   # labs, fresh from the month-3 drop
```

## Where you are

- A folder workflow replaces the one-table-at-a-time pattern.
- Four tables are versioned; the project has a real shape.
- `datom_status()` and `datom_validate()` are your two go-to consistency
  checks.

This is the end of the local Get Started journey. From here,
[Starting on S3](start-on-s3.html) shows the same workflow with data in object
storage, and [Tracing Data Lineage](source-lineage.html) shows how derived
tables record where they came from.

## Tear down

When you're done, remove the project. Pick the one option that fits, and copy
only that chunk:

```{r}
# Option A -- full scripted teardown (deletes local files AND the GitHub repo).
# Do this BEFORE any manual unlink().
datom_repo_delete(conn, confirm = "STUDY_001")
```

```{r}
# Option B -- local only (the GitHub repo stays; delete it from the UI later).
unlink(c(dev_dir, data_dir), recursive = TRUE)
```

Do not call `unlink()` before `datom_repo_delete()` -- removing the local
clone first strips the GitHub remote reference and the remote repo will not
be deleted.
