---
title: "Month 2 Arrives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Month 2 Arrives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

> **Where we left off:** STUDY-001 is initialized. One table (`dm`) holds the
> first 4 subjects. The data git repo is live on GitHub.

A month has passed. Twelve more subjects have enrolled. The data management
team sends you the month-2 extract, and you need to register the new state
without losing the old one.

This article introduces three capabilities: writing a second version of an
existing table, asking datom for a table's full history, and reading any
historical version by its SHA.

## Reconnect (only if you closed your session)

If you're continuing in the same R session as
[First Extract](first-extract.html), you already have `conn` and `store` --
skip ahead to the extract below.

If you closed your session, re-run the same configuration block to reconnect.
The project already exists, so you reconnect with `datom_get_conn()` rather
than initializing again:

```{r}
library(datom)
library(fs)

# The same machine-specific settings you used in article 1.
project_name <- "STUDY_001"
dev_dir      <- path(tempdir(), "study_001_dev")
data_dir     <- path(tempdir(), "study_001_data")
github_pat   <- keyring::key_get("GITHUB_PAT")

store <- datom_store(
  governance = NULL,
  data       = datom_store_local(path = data_dir),
  github_pat = github_pat
)

conn <- datom_get_conn(path = dev_dir, store = store)
```

Temp directories do not survive a new R session. To follow the journey across
sessions, point `dev_dir` and `data_dir` at persistent paths in article 1.

## The month-2 extract

```{r}
library(datom)

dm_m2 <- datom_example_data("dm", cutoff_date = "2026-02-28")
nrow(dm_m2)
#> [1] 16
```

Twelve new subjects have been added. The month-1 rows are still there,
unchanged.

## Write the new version

```{r}
datom_write(
  conn,
  data    = dm_m2,
  name    = "dm",
  message = "DM extract through 2026-02-28"
)
#> v Wrote "dm" (full): "5c1a3f7b"
```

`(full)` means both the data and the metadata changed -- datom uploaded a new
parquet file and recorded a new entry in the version history. The previous
version is **still on disk and still in git**; nothing was overwritten.

## Re-running is a no-op

datom is content-addressed: if you call `datom_write()` again with the same
data, it detects that nothing has changed and returns without producing a
new version.

```{r}
datom_write(conn, data = dm_m2, name = "dm")
#> i No changes detected for "dm". Skipping write.
```

This is what makes pipelines safe to re-run -- accidental duplicate writes
cost nothing and pollute no history.

## Look at the history

```{r}
datom_history(conn, "dm")
#>    version  data_sha timestamp            author commit_message
#> 1 5c1a3f7b 9e8f1c2d 2026-02-28T10:14:02Z you    DM extract through 2026-02-28
#> 2 a8ee7a31 4b6d0a7e 2026-01-28T09:02:11Z you    Initial DM extract
```

Two rows, newest first. The `version` column is the **metadata SHA** -- the
identifier you use to pin a read.

## Pin a read to a specific version

The default `datom_read()` returns the current version:

```{r}
dm_now <- datom_read(conn, "dm")
nrow(dm_now)
#> [1] 16
```

Pass a `version` to retrieve any historical snapshot:

```{r}
hist   <- datom_history(conn, "dm")
m1_ver <- hist$version[nrow(hist)]   # oldest row is the month-1 version

dm_m1_again <- datom_read(conn, "dm", version = m1_ver)
nrow(dm_m1_again)
#> [1] 4
```

The data you get back is bit-for-bit identical to what you wrote in
article 1, even though the current version of `dm` has 16 rows. This is the
property your future statisticians, regulators, and auditors rely on.

## The analyst's view (reader role)

A statistician doesn't need your developer connection or local clone. They
build a **reader** connection -- data store only, no PAT -- and pin the same
SHA:

```{r}
reader_store <- datom_store(
  governance = NULL,
  data       = datom_store_local(path = data_dir)
)
reader_conn <- datom_get_conn(store = reader_store, project_name = project_name)

nrow(datom_read(reader_conn, "dm", version = m1_ver))
#> [1] 4
```

Same SHA, same bytes -- whether read by the engineer who wrote it or an
analyst on another machine.

## Where you are

- `dm` has two versions; you can read either at any time.
- Re-writing the same data is a free no-op.
- Version SHAs are stable across machines -- share one and a colleague reads
  the same bytes you did.

**Carry on:** keep this R session open and reuse `conn`, `store`, and your
settings in the next article. Do **not** tear down -- the journey continues.
In [A Folder of Extracts](folder-of-extracts.html), a whole folder arrives
instead of a single table, and we use `datom_sync()` to import them at once.
