---
title: "S3 Setup and Promotion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{S3 Setup and Promotion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

S3 is where datom turns into shared, governed infrastructure. By the end of
this article you can:

- Give a teammate read-only access to the project -- your cloud admin
  generates scoped credentials, the teammate stores them in keyring, and
  datom's reader role works out of the box.
- Register the project in a shared governance portfolio, discoverable by
  name across your organization.

Moving to S3 also lays the foundation for capabilities datom will integrate
over time: access logs, automated retention rules, cross-account replication.
None of that requires any code changes on your part -- it follows from
choosing object storage as the backend.

From your code, nothing changes: the same `datom_write()`, `datom_read()`,
and `datom_history()` calls you've used so far.

## Two ways to read this article

- **Starting fresh on S3** — skip the migration note and the promotion
  steps; jump straight to [Set up AWS credentials](#set-up-aws-credentials).
- **Promoting an existing local-filesystem project to S3** — follow in
  order. You'll snapshot the current data, retire the local project, then
  re-establish it on S3.

Both paths converge at [Build the S3 store](#build-the-s3-store).

> **Where we left off (promotion path):** STUDY-001 has four tables
> (`dm`, `ex`, `lb`, `ae`), all in a local filesystem store. The data git
> repo is on GitHub. No governance layer is attached yet -- articles 1-3
> stayed deliberately solo.

## What promotion looks like today (promotion path only)

> **Starting fresh on S3?** Skip this section.

A built-in, history-preserving migration (`datom_migrate_data()`) is
planned but not yet shipped. Today, promoting a project means:

1. Snapshot the current version of each table.
2. Retire the local project (`datom_decommission()`).
3. Initialize a new project on S3 with the same name.
4. Re-write the snapshotted tables as version 1 on S3.

The trade-off is that **per-table version history from the local era is
not carried forward** — only the latest version of each table is. For a
study with a few months of extracts this is cheap; the git commit log
preserves the narrative even when the data history restarts.

If preserving full per-version history across the move matters to you right
now, the cleaner path is to start a new project directly on S3 (next
section) and write your data there going forward.

## Set up AWS credentials

datom uses the [keyring](https://r-lib.github.io/keyring/) package for
credentials, the same way it stored your GitHub PAT in article 1. Set your
AWS access keys once:

```{r}
keyring::key_set("AWS_ACCESS_KEY_ID")
keyring::key_set("AWS_SECRET_ACCESS_KEY")
```

If your organization issues short-lived session tokens (STS), set the
session token too:

```{r}
keyring::key_set("AWS_SESSION_TOKEN")
```

You will also need:

- A **bucket** you can read and write. datom does not create buckets for
  you — bucket lifecycle (encryption, versioning, retention) is your
  organization's policy domain, not datom's.
- A **prefix** within the bucket if you share the bucket with other
  projects. Below we use `study-001/`.

The full credential reference, including how to scope a reader-only PAT and
how to handle SSO + assume-role flows, is in
[Credentials in Practice](credentials-in-practice.html).

## Resume the prior state (promotion path only)

```{r}
state <- source(
  system.file("vignette-setup", "resume_article_4.R", package = "datom")
)$value

old_conn  <- state$conn
dev_dir   <- state$dev_dir
```

`old_conn` is the local-backend conn from article 3. We'll use it to read
the four current tables, then decommission it.

## Snapshot the current data (promotion path only)

Before tearing anything down, capture the latest state of each table in
memory:

```{r}
library(datom)

cutoff   <- "2026-03-28"
snapshot <- list(
  dm = datom_read(old_conn, "dm"),
  ex = datom_read(old_conn, "ex"),
  lb = datom_read(old_conn, "lb"),
  ae = datom_read(old_conn, "ae")
)
```

Each element is the current version's data frame.

## Decommission the local project (promotion path only)

`datom_decommission()` deletes the GitHub repo and clears the local clone
and parquet store. It is **destructive** and requires you to type the
project name as `confirm` to proceed.

```{r}
datom_decommission(old_conn, confirm = "STUDY_001")
#> i Removing data store contents under datom/STUDY_001/
#> v Deleted GitHub repo `study-001-data`
#> v Removed local data clone /tmp/.../study_001_data
```

Because no governance layer was attached, decommission is purely
data-side: no registry to unregister from, no gov storage to clean up.

## Build the S3 store

```{r}
library(fs)

dev_dir <- path(tempdir(), "study_001_dev")  # fresh clone target

aws_data <- datom_store_s3(
  bucket     = "your-org-datom-data",   # <-- replace with a bucket you own
  prefix     = "study-001/",
  region     = "us-east-1",
  access_key = keyring::key_get("AWS_ACCESS_KEY_ID"),
  secret_key = keyring::key_get("AWS_SECRET_ACCESS_KEY")
)

store <- datom_store(
  governance = NULL,                                  # still no gov; attached below
  data       = aws_data,
  github_pat = keyring::key_get("GITHUB_PAT")
)
```

## Initialize STUDY_001 on S3

```{r}
datom_init_repo(
  path         = dev_dir,
  store        = store,
  create_repo  = TRUE,
  repo_name    = "study-001-data"
)

conn <- datom_get_conn(path = dev_dir, store = store)
print(conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "developer"
#> * Backend: "s3"
#> * Root: "your-org-datom-data"
#> * Prefix: "study-001/"
#> * Governance: not attached
```

The data backend is now `"s3"`. From here on, every `datom_write()` uploads
parquet to S3 and every `datom_read()` streams it back from S3.

## Attach the governance layer

Now that STUDY-001 lives somewhere a teammate can reach, register it in a
shared governance repo. `datom_attach_gov()` creates the gov GitHub repo
(once per organization), records STUDY-001's data location in
`projects/STUDY_001/ref.json`, and updates the data repo's `project.yaml`
so any future conn from this clone knows where gov lives.

```{r}
gov_store <- datom_store_s3(
  bucket     = "your-org-datom-gov",    # <-- a separate gov bucket
  prefix     = "datom-gov/",
  region     = "us-east-1",
  access_key = keyring::key_get("AWS_ACCESS_KEY_ID"),
  secret_key = keyring::key_get("AWS_SECRET_ACCESS_KEY")
)

conn <- datom_attach_gov(
  conn        = conn,
  gov_store   = gov_store,
  create_repo = TRUE,
  repo_name   = "datom-governance"
)
#> v Created gov GitHub repo `datom-governance`
#> v Cloned gov repo to /tmp/.../datom-governance
#> v Registered STUDY_001 in governance
#> v Updated project.yaml with governance pointer

print(conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "developer"
#> * Backend: "s3"
#> * Root: "your-org-datom-data"
#> * Prefix: "study-001/"
#> * Gov backend: "s3"
#> * Gov root: "your-org-datom-gov"
```

Once attached, gov cannot be detached — `project.yaml`'s
`storage.governance` block is permanent. Subsequent projects in the same
organization reuse the same gov repo and bucket; you only run
`create_repo = TRUE` once.

## Replay the four tables

```{r}
datom_write(conn, snapshot$dm, "dm",
            message = "Re-establish dm on S3 (was local through 2026-03-28)")
datom_write(conn, snapshot$ex, "ex",
            message = "Re-establish ex on S3 (was local through 2026-03-28)")
datom_write(conn, snapshot$lb, "lb",
            message = "Re-establish lb on S3 (was local through 2026-03-28)")
datom_write(conn, snapshot$ae, "ae",
            message = "Re-establish ae on S3 (was local through 2026-03-28)")
```

Each table now has version 1 in the S3-backed project. The history of
versions 1–3 from the local project is gone; the **commit messages** above
are your audit trail for that.

## Confirm

```{r}
datom_list(conn)
#>   name current_version current_data_sha last_updated
#> 1   ae    19f44e3a       e91d04ff         2026-04-29T...
#> 2   dm    8a3b21cc       c2e80a14         2026-04-29T...
#> 3   ex    5d72e0f1       88a73e02         2026-04-29T...
#> 4   lb    c1ffea90       4c3812dd         2026-04-29T...

# Note: data_sha values match the local project (parquet bytes are identical).
# Version SHAs are new because metadata (commit times, messages) differs.
```

The `data_sha` column matches the local project's last `data_sha` for each
table — the **bytes** are identical, datom is just storing them in a
different place. Version SHAs are different because they include metadata
that changed (timestamps, commit messages).

## Where you are

- STUDY_001 lives on S3. Your local clone is just a working copy of the git
  metadata.
- Governance is attached: STUDY_001 is registered in a shared gov repo and
  gov bucket. Future projects in this organization reuse both.
- The four tables are at version 1 again, with honest commit messages
  recording the local-era history.
- `ref.json` in the gov repo points at the S3 bucket; any teammate who
  clones the gov repo and has S3 read access can discover and read the
  data.

In the next article, you **hand the project off to a statistician** who
needs to read the data without write access — the canonical reader role.

````
