---
title: "Promoting to S3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Promoting to S3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

> **Where we left off:** STUDY-001 has four tables (`dm`, `ex`, `lb`, `ae`),
> all on your laptop. The data git repo is on GitHub. No governance layer
> is attached yet — articles 1-3 stayed deliberately solo.

A colleague on a different laptop now needs to read the data. Local
filesystem storage was fine when you were the only one writing; it stops
working the moment a second person needs the same bytes.

This article does two things, both triggered by the same moment ("someone
else needs to read this"):

1. **Promote** STUDY-001 from local-filesystem data storage to **S3**.
2. **Attach** the governance layer with `datom_attach_gov()` — a shared
   registry across projects, used by readers to discover where data lives
   and by org-wide tooling to see the portfolio. Gov is opt-in and only
   pays for itself once a project is shared.

After this article, the data git repo on GitHub is unchanged, the parquet
bytes live in an S3 bucket, and STUDY-001 is registered in a governance
repo any teammate can clone to find it.

## A note on local -> S3 migration

datom's `ref.json` has slots for `previous` data locations and a
`migration_history.json` is initialized for every project. The intent is a
future `datom_migrate_data()` that copies bytes from one store to another
without resetting history. **That capability isn't shipped yet** (planned
for a future release). Until it is, the practical path is to retire the
local project and re-establish it on S3.

We choose this honest, slightly-lossy path here because:

- The history of a *brand-new* study with three months of extracts is
  recoverable: re-write the four current tables and you're done.
- Studies that have run for a year on local storage shouldn't have run on
  local storage — they are the audience for `datom_migrate_data()` later.

If preserving history across a backend change matters to your project
**right now**, start on S3 from article 1 and skip this article.

## Set up AWS credentials

datom uses the [keyring](https://r-lib.github.io/keyring/) package for
credentials, the same way it stored your GitHub PAT in article 1. Set your
AWS access keys once:

```{r}
keyring::key_set("AWS_ACCESS_KEY_ID")
keyring::key_set("AWS_SECRET_ACCESS_KEY")
```

If your organization issues short-lived session tokens (STS), set the
session token too:

```{r}
keyring::key_set("AWS_SESSION_TOKEN")
```

You will also need:

- A **bucket** you can read and write. datom does not create buckets for
  you — bucket lifecycle (encryption, versioning, retention) is your
  organization's policy domain, not datom's.
- A **prefix** within the bucket if you share the bucket with other
  projects. Below we use `study-001/`.

The full credential reference, including how to scope a reader-only PAT and
how to handle SSO + assume-role flows, is in
[Credentials in Practice](credentials-in-practice.html).

## Resume the prior state

```{r}
state <- source(
  system.file("vignette-setup", "resume_article_4.R", package = "datom")
)$value

old_conn  <- state$conn
study_dir <- state$study_dir
```

`old_conn` is the local-backend conn from article 3. We'll use it to read
the four current tables, then decommission it.

## Snapshot the current data

Before tearing anything down, capture the latest state of each table in
memory:

```{r}
library(datom)

cutoff   <- "2026-03-28"
snapshot <- list(
  dm = datom_read(old_conn, "dm"),
  ex = datom_read(old_conn, "ex"),
  lb = datom_read(old_conn, "lb"),
  ae = datom_read(old_conn, "ae")
)
```

Each element is the current version's data frame.

## Decommission the local project

`datom_decommission()` deletes the GitHub repo and clears the local clone
and parquet store. It is **destructive** and requires you to type the
project name as `confirm` to proceed.

```{r}
datom_decommission(old_conn, confirm = "STUDY_001")
#> i Removing data store contents under datom/STUDY_001/
#> v Deleted GitHub repo `study-001-data`
#> v Removed local data clone /tmp/.../study_001_data
```

Because no governance layer was attached, decommission is purely
data-side: no registry to unregister from, no gov storage to clean up.

## Build the S3 store

```{r}
library(fs)

study_dir <- path(tempdir(), "study_001_data")  # fresh clone target

aws_data <- datom_store_s3(
  bucket     = "your-org-datom-data",   # <-- replace with a bucket you own
  prefix     = "study-001/",
  region     = "us-east-1",
  access_key = keyring::key_get("AWS_ACCESS_KEY_ID"),
  secret_key = keyring::key_get("AWS_SECRET_ACCESS_KEY")
)

store <- datom_store(
  governance = NULL,                                  # still no gov; attached below
  data       = aws_data,
  github_pat = keyring::key_get("GITHUB_PAT")
)
```

## Re-initialize STUDY_001 on S3

```{r}
datom_init_repo(
  path         = study_dir,
  project_name = "STUDY_001",
  store        = store,
  create_repo  = TRUE,
  repo_name    = "study-001-data"
)

conn <- datom_get_conn(path = study_dir, store = store)
print(conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "developer"
#> * Backend: "s3"
#> * Root: "your-org-datom-data"
#> * Prefix: "study-001/"
#> * Governance: not attached
```

The data backend is now `"s3"`. From here on, every `datom_write()` uploads
parquet to S3 and every `datom_read()` streams it back from S3.

## Attach the governance layer

Now that STUDY-001 lives somewhere a teammate can reach, register it in a
shared governance repo. `datom_attach_gov()` creates the gov GitHub repo
(once per organization), records STUDY-001's data location in
`projects/STUDY_001/ref.json`, and updates the data repo's `project.yaml`
so any future conn from this clone knows where gov lives.

```{r}
gov_store <- datom_store_s3(
  bucket     = "your-org-datom-gov",    # <-- a separate gov bucket
  prefix     = "datom-gov/",
  region     = "us-east-1",
  access_key = keyring::key_get("AWS_ACCESS_KEY_ID"),
  secret_key = keyring::key_get("AWS_SECRET_ACCESS_KEY")
)

conn <- datom_attach_gov(
  conn        = conn,
  gov_store   = gov_store,
  create_repo = TRUE,
  repo_name   = "datom-governance"
)
#> v Created gov GitHub repo `datom-governance`
#> v Cloned gov repo to /tmp/.../datom-governance
#> v Registered STUDY_001 in governance
#> v Updated project.yaml with governance pointer

print(conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "developer"
#> * Backend: "s3"
#> * Root: "your-org-datom-data"
#> * Prefix: "study-001/"
#> * Gov backend: "s3"
#> * Gov root: "your-org-datom-gov"
```

Once attached, gov cannot be detached — `project.yaml`'s
`storage.governance` block is permanent. Subsequent projects in the same
organization reuse the same gov repo and bucket; you only run
`create_repo = TRUE` once.

## Replay the four tables

```{r}
datom_write(conn, snapshot$dm, "dm",
            message = "Re-establish dm on S3 (was local through 2026-03-28)")
datom_write(conn, snapshot$ex, "ex",
            message = "Re-establish ex on S3 (was local through 2026-03-28)")
datom_write(conn, snapshot$lb, "lb",
            message = "Re-establish lb on S3 (was local through 2026-03-28)")
datom_write(conn, snapshot$ae, "ae",
            message = "Re-establish ae on S3 (was local through 2026-03-28)")
```

Each table now has version 1 in the S3-backed project. The history of
versions 1–3 from the local project is gone; the **commit messages** above
are your audit trail for that.

## Confirm

```{r}
datom_list(conn)
#>   name current_version current_data_sha last_updated
#> 1   ae    19f44e3a       e91d04ff         2026-04-29T...
#> 2   dm    8a3b21cc       c2e80a14         2026-04-29T...
#> 3   ex    5d72e0f1       88a73e02         2026-04-29T...
#> 4   lb    c1ffea90       4c3812dd         2026-04-29T...

# Note: data_sha values match the local project (parquet bytes are identical).
# Version SHAs are new because metadata (commit times, messages) differs.
```

The `data_sha` column matches the local project's last `data_sha` for each
table — the **bytes** are identical, datom is just storing them in a
different place. Version SHAs are different because they include metadata
that changed (timestamps, commit messages).

## Where you are

- STUDY_001 lives on S3. Your local clone is just a working copy of the git
  metadata.
- Governance is attached: STUDY_001 is registered in a shared gov repo and
  gov bucket. Future projects in this organization reuse both.
- The four tables are at version 1 again, with honest commit messages
  recording the local-era history.
- `ref.json` in the gov repo points at the S3 bucket; any teammate who
  clones the gov repo and has S3 read access can discover and read the
  data.

In the next article, you **hand the project off to a statistician** who
needs to read the data without write access — the canonical reader role.

````
