---
title: "A Second Engineer Joins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Second Engineer Joins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

> **Where we left off:** STUDY-001 is on S3. The statistician can read.

The study has grown enough that you bring in a second data engineer. Unlike
the statistician, the second engineer needs to **write** — to register the
month-4 lab extract while you're on vacation.

This article introduces `datom_clone()` (the way a new engineer enters an
existing project) and demonstrates the **pull-before-push** discipline that
keeps two engineers from clobbering each other.

This article assumes you've completed Article 4. Resume the prior state if
needed:

```{r}
state <- source(
  system.file("vignette-setup", "resume_article_6.R", package = "datom")
)$value

conn      <- state$conn         # original engineer
study_dir <- state$study_dir
```

## The second engineer clones

The second engineer (call her Bea) sets up her credentials the same way you
did:

```{r}
keyring::key_set("GITHUB_PAT")
keyring::key_set("AWS_ACCESS_KEY_ID")
keyring::key_set("AWS_SECRET_ACCESS_KEY")
```

Bea then clones into a fresh directory on her laptop. She uses the engineer
store (read+write):

```{r}
library(datom)
library(fs)

bea_dir <- path(tempdir(), "bea_study_001_data")  # different from yours

bea_store <- datom_store(
  governance     = state$gov_component,    # same gov as you
  data           = state$data_s3,          # same bucket as you
  github_pat     = keyring::key_get("GITHUB_PAT"),
  data_repo_url  = state$data_repo_url,
  gov_repo_url   = state$gov_repo_url,
  gov_local_path = path(tempdir(), "bea_gov_clone")
)

bea_conn <- datom_clone(path = bea_dir, store = bea_store)
print(bea_conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Role: "developer"
#> * Backend: "s3"
```

`datom_clone()` does three things in order: clones the data git repo into
`bea_dir`, clones the gov repo (if not already present), and returns a
developer conn. Bea now sees the same four tables you do.

```{r}
datom_list(bea_conn)
#>   name current_version current_data_sha last_updated
#> 1   ae    19f44e3a       e91d04ff        2026-04-29T...
#> 2   dm    8a3b21cc       c2e80a14        2026-04-29T...
#> 3   ex    5d72e0f1       88a73e02        2026-04-29T...
#> 4   lb    c1ffea90       4c3812dd        2026-04-29T...
```

## Concurrent writes

You and Bea both have month-4 extracts in your inboxes. You write `lb`:

```{r}
# In your session
lb_m4 <- datom_example_data("lb", cutoff_date = "2026-04-28")
datom_write(conn, lb_m4, "lb",
            message = "lb extract through 2026-04-28")
#> v Wrote "lb" (full): "..."
```

At the same time, Bea writes `ae` from her clone:

```{r}
# In Bea's session
ae_m4 <- datom_example_data("ae", cutoff_date = "2026-04-28")
datom_write(bea_conn, ae_m4, "ae",
            message = "ae extract through 2026-04-28")
```

What happens to Bea's write? It depends on which one of you committed
first. The second commit **fails on push**:

```{r}
#> x Failed to push to origin/main: rejected (non-fast-forward)
#> i Run datom_pull(conn) to incorporate origin's changes, then retry the write.
```

This is the same conflict signature any git user has seen. datom's response
is the standard one: pull, then retry.

## Pull, then retry

Bea runs `datom_pull()` to bring her clone up to date with your `lb` write:

```{r}
datom_pull(bea_conn)
#> i Fetching from origin...
#> v 1 commit pulled.
#> v Manifest updated: lb -> new version
```

Now Bea retries her `ae` write. Because `ae` is a different table, there's
no conflict at the data level — the two writes commute, and after the pull
they layer cleanly:

```{r}
datom_write(bea_conn, ae_m4, "ae",
            message = "ae extract through 2026-04-28")
#> v Wrote "ae" (full): "..."
```

If Bea had been writing `lb` (the same table you wrote), her retry would
produce a **third** version of `lb` — datom does not have a same-table
merge concept, because there's no semantic merge of two parquet files. Each
`datom_write()` is a snapshot; the version order tells you who wrote when.

## Push discipline

The rule for two-engineer projects:

1. **Pull before each writing session**, not just after a conflict. A clean
   pull at the top of the day prevents most conflicts.
2. **Pull again before push** if the writing session was long. Conflicts
   detected at push time are the more annoying kind.
3. **Communicate which tables you're touching** when working in parallel.
   datom doesn't have row-level locks; it's still a human coordination
   problem, just one that's now mechanically enforced.

## Where you are

- Two engineers, one project, on the same S3 store.
- Concurrent writes to **different tables** are conflict-free after a pull.
- Concurrent writes to the **same table** produce sequential versions,
  ordered by who pushed first.
- Bea has the same capabilities you do — there is no "primary" engineer;
  the project is shared.

The user-journey track continues with the manager view in **Governing a
Study Portfolio**, where STUDY-002 joins your gov repo and you start
caring about the registry shape across projects.

````
