---
title: "Tracing Data Lineage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracing Data Lineage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

datom records two complementary lineage fields in every table's metadata:

- `parents` -- the **immediate** inputs used to derive this table
  (identified by `metadata_sha`, the datom version identifier).
- `source_lineage` -- the **transitive closure** of all raw tables that
  contributed data, stored as a flat list of `{project, table,
  version_sha}` entries where `version_sha` is the content SHA of the
  source parquet file.

Together they answer two different audit questions without walking a DAG:

| Question | Field | Tool |
|---|---|---|
| "What exact versions did my script read?" | `parents` | `datom_get_lineage(depth = "parents")` |
| "What raw data is ultimately in this table?" | `source_lineage` | `datom_get_lineage(depth = "source")` |

This article shows how lineage is recorded and how to query and validate it.

---

## Setup

```{r}
library(datom)
```

The code below uses a local-backend sandbox so it runs without S3 or
GitHub credentials. In production, replace `datom_store_local()` with
`datom_store_s3()` and `datom_store()` -- the API is identical.

```{r}
# Not evaluated -- illustrates a real-world setup.
# In practice this is done once per project with datom_init_repo().
store <- datom_store(
  data      = datom_store_s3(bucket = "my-datom-bucket", prefix = "lineage-demo"),
  gov       = datom_store_s3(bucket = "my-gov-bucket",   prefix = "gov")
)
```

---

## A simple study: raw imports to derived analysis

Imagine a clinical study with two raw source tables:

```{r}
raw_dm <- datom_example_data()          # Demographics
raw_lb <- datom_example_data()          # Lab results (illustrative)
```

They are imported via `datom_sync()`. datom automatically records each
imported table's own SHA as its `source_lineage` -- a single self-entry.

```{r}
datom_sync(conn, list(dm = raw_dm, lb = raw_lb))
```

After syncing, the imported tables carry their own lineage:

```{r}
datom_get_lineage(conn, "dm", depth = "source")
#> [[1]]
#> [[1]]$project
#> [1] "my-study"
#>
#> [[1]]$table
#> [1] "dm"
#>
#> [[1]]$version_sha
#> [1] "abc123..."
```

The self-entry means: "the raw content of `dm` is itself a source."

---

## Deriving a table

A downstream script reads both raw tables and produces a cleaned
demographics table:

```{r}
# Read the current versions
raw_dm_data <- datom_read(conn, "dm")
raw_lb_data <- datom_read(conn, "lb")

# ... cleaning logic ...
dm_clean <- raw_dm_data  # simplified

# Retrieve the metadata shas to identify exactly which versions were read
dm_version  <- datom_history(conn, "dm")[1, "version"]
lb_version  <- datom_history(conn, "lb")[1, "version"]
```

When writing the derived table, supply `parents` (which versions were
read) and `source_lineage` (the union of all raw sources):

```{r}
# datom_get_lineage fetches the source_lineage of each parent
dm_lineage <- datom_get_lineage(conn, "dm", depth = "source")
lb_lineage <- datom_get_lineage(conn, "lb", depth = "source")

# Union is a simple c() here because both are single self-entries;
# dpbuild automates this for complex pipelines.
full_lineage <- c(dm_lineage, lb_lineage)

datom_write(
  conn,
  data           = dm_clean,
  name           = "dm_clean",
  parents        = list(
    list(source = "my-study", table = "dm", version = dm_version),
    list(source = "my-study", table = "lb", version = lb_version)
  ),
  source_lineage = full_lineage
)
```

Now `dm_clean` knows which raw tables it came from:

```{r}
datom_get_lineage(conn, "dm_clean", depth = "source")
#> [[1]]
#> $project
#> [1] "my-study"
#> $table
#> [1] "dm"
#> $version_sha
#> [1] "abc123..."
#>
#> [[2]]
#> $project
#> [1] "my-study"
#> $table
#> [1] "lb"
#> $version_sha
#> [1] "def456..."
```

And which immediate versions it was derived from:

```{r}
datom_get_lineage(conn, "dm_clean", depth = "parents")
#> [[1]]
#> $source
#> [1] "my-study"
#> $table
#> [1] "dm"
#> $version
#> [1] "..."   # metadata_sha of dm at derivation time
```

---

## Propagating lineage further downstream

An analysis table derived from `dm_clean` propagates the lineage by
unioning `dm_clean`'s `source_lineage` with any additional raw inputs.
Since `dm_clean` already encodes both `dm` and `lb`, the analysis table
inherits the full picture without re-reading the original files:

```{r}
# One hop: fetch dm_clean's source lineage (already transitive)
clean_lineage <- datom_get_lineage(conn, "dm_clean", depth = "source")

datom_write(
  conn,
  data           = analysis_pop,
  name           = "analysis_pop",
  parents        = list(
    list(source = "my-study", table = "dm_clean", version = clean_version)
  ),
  source_lineage = clean_lineage  # no new raw inputs; pass through
)
```

---

## Validating lineage consistency

`datom_validate_lineage()` checks whether a table's declared
`source_lineage` is consistent with its parents' lineages:

```{r}
datom_validate_lineage(conn, "dm_clean")
#> v Lineage OK for "dm_clean" (2 source entries, 2 parents checked).
#> $status
#> [1] "ok"
#> $missing
#> list()
#> $extra
#> list()
#> $wrong_version
#> list()
#> $message
#> [1] "ok: 2 source entries match computed lineage from 2 parent(s)"
```

To see what a mismatch looks like, imagine `dm_clean` was written with an
incorrect `version_sha` for the `lb` source. The validator catches it:

```{r}
datom_validate_lineage(conn, "dm_clean_bad")
#> ! Lineage mismatch for "dm_clean_bad".
#> $status
#> [1] "mismatch"
#> $missing
#> list()
#> $extra
#> list()
#> $wrong_version
#> [[1]]
#>   project table version_sha_declared version_sha_computed
#>   <chr>   <chr> <chr>                <chr>
#> 1 my-study lb   wrong000...          def456...
```

---

## Key points

- `datom_sync()` auto-populates `source_lineage` for imported tables
  (a single self-entry using the file's content SHA).
- `datom_write()` requires `source_lineage` whenever `parents` is
  supplied -- structural mandate to prevent silent omissions.
- `datom_get_lineage()` is a single-read operation -- no DAG traversal,
  no recursive network calls.
- `datom_validate_lineage()` returns a structured result; it does not
  hard-abort, so callers can embed it in CI pipelines and decide on
  action.
- **Walker invariant**: `source_lineage` entries are terminal leaves.
  Lineage walkers must follow `parents`, never `source_lineage`.
