---
title: "Starting on S3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Starting on S3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

```{=html}
<link rel="stylesheet" href="../inst/vignette-setup/datom.css">
```

<div class="datom-callout datom-callout-goal" markdown="1">
**Goal:** Stand up a versioned datom project whose data lives directly in
**Amazon S3** -- no local-filesystem detour and no migration step. Write your
first table, read it back, and confirm git is tracking the history. This is the
"I'm already sold, start me in object storage" path.
</div>

> **Want to dabble locally first?** Start with
> [First Extract](first-extract.html) instead -- it uses a local filesystem
> store and needs no AWS account. The two paths use the **same functions**; the
> only difference is which store you build. You can read either one first.

You are the data engineer for **STUDY-001**, a Phase II clinical trial, and your
team already works in S3. There is no reason to stage data on a laptop: you want
the very first extract to land in the shared object store, versioned from commit
one. datom supports that directly. Metadata still lives in a git repository (so
version history is diff-able and reproducible across machines); only the choice
of data store changes.

## Requirements

datom keeps **metadata in git** and **data wherever you tell it to live**. For
this article that is an S3 bucket. You need three things:

- **A GitHub account** with a personal access token (PAT) scoped to `repo`.
  Store it in your OS keychain once with `keyring::key_set("GITHUB_PAT")`. The
  `gh` CLI is **not** required -- datom creates the metadata repo through the
  GitHub REST API directly using your PAT.
- **An S3 bucket** you can read and write. datom does **not** create buckets:
  bucket lifecycle (encryption, versioning, retention) is your organization's
  policy domain. A common convention is one bucket per study, with raw data at
  the bucket root (an empty prefix) and derived products under named prefixes
  (`adam/`, `tlf/`).
- **AWS credentials** (an access key and secret key) authorized for that bucket.

<div class="datom-callout datom-callout-tip" markdown="1">
**Verify your keyring setup** before continuing:

```r
nzchar(keyring::key_get("GITHUB_PAT"))   # should return TRUE
```
</div>

## Configure your environment

Everything machine-specific lives here. Credentials have several valid sources,
so they get their own section (and their own chunks) below; set the rest once:

```{r}
library(datom)
library(fs)

# --- Settings you control --------------------------------------------------
project_name <- "STUDY_001"        # logical project name (recorded in metadata)
repo_name    <- "study-001-data"   # GitHub repo name for the metadata repo

bucket <- "study-001-data"         # an S3 bucket you can read/write
                                   #   (datom does NOT create buckets)
prefix <- NULL                     # raw data at the bucket root; use e.g.
                                   #   "adam/" for a derived-products prefix
region <- "us-east-1"              # the bucket's AWS region

# Local working directory for the metadata git clone. The data itself never
# lands here -- it goes straight to S3. A temp dir is fine for this walkthrough.
dev_dir <- path(tempdir(), "study_001_dev")

# GitHub PAT (scoped to `repo`), read from your OS keychain by name.
github_pat <- keyring::key_get("GITHUB_PAT")
# ---------------------------------------------------------------------------
```

## Set your AWS credentials

`datom_store_s3()` takes `access_key` and `secret_key` as plain strings. datom
never reads them from the environment on your behalf -- you pass the values in
explicitly. **Run exactly one** of the chunks below, whichever fits your
environment, then continue.

```{r}
# Option A -- keyring (recommended for an interactive developer machine)
access_key <- keyring::key_get("AWS_ACCESS_KEY_ID")
secret_key <- keyring::key_get("AWS_SECRET_ACCESS_KEY")
```

```{r}
# Option B -- environment variables (CI/CD, Docker)
access_key <- Sys.getenv("AWS_ACCESS_KEY_ID")
secret_key <- Sys.getenv("AWS_SECRET_ACCESS_KEY")
```

```{r}
# Option C -- inline (fine for a quick session; never commit these values)
access_key <- "AKIAIOSFODNN7EXAMPLE"
secret_key <- "wJalrXUtnFEMI/K7MDENG/bPxRfiCYEXAMPLEKEY"
```

## Build the S3 store

A **store** bundles the addresses datom needs: where parquet bytes go and the
GitHub PAT that lets datom push metadata. The data component is now an S3
component instead of a local one.

Governance is **not attached** (`governance = NULL`). A solo project's location
authority is its own `project.yaml`; you do not need a portfolio register to get
started. Governance is opt-in and added later (see
[Governance and migration come later](#governance-and-migration-come-later)).

```{r}
data_component <- datom_store_s3(
  bucket     = bucket,
  prefix     = prefix,
  region     = region,
  access_key = access_key,
  secret_key = secret_key
)

store <- datom_store(
  governance = NULL,
  data       = data_component,
  github_pat = github_pat
)
```

By default `datom_store_s3()` validates connectivity to the bucket as the store
is built, so a credential or permission problem surfaces immediately rather than
at first write.

## Initialize the data repository

The local working directory (`dev_dir`) holds only the **git clone** -- the
metadata repository. The data itself never lands there; it goes straight to S3.

```{r}
datom_init_repo(
  path         = dev_dir,
  project_name = project_name,
  store        = store,
  create_repo  = TRUE,
  repo_name    = repo_name
)
```

This creates a GitHub repo, clones it into `dev_dir`, and commits a
`project.yaml` that records the project's S3 data store address. No parquet data
is pushed to GitHub -- only metadata commits travel over the wire.

## Connect

```{r}
conn <- datom_get_conn(path = dev_dir, store = store)
print(conn)
#> -- datom connection
#> * Project: "STUDY_001"
#> * Backend: "s3"
#> * Role: "developer"
#> * Data root: "study-001-data"
#> * Data region: "us-east-1"
#> * Governance: not attached
#> * Path: "/tmp/.../study_001_dev"
#> * Data repo: <https://github.com/.../study-001-data>
```

## Write your first extract

The month-1 extract has just landed. Load the demographics snapshot for
subjects enrolled by 2026-01-28 and write it as a versioned datom table:

```{r}
dm_m1 <- datom_example_data("dm", cutoff_date = "2026-01-28")

datom_write(
  conn,
  data    = dm_m1,
  name    = "dm",
  message = "Initial DM extract through 2026-01-28"
)
#> v Wrote "dm" (full): "a8ee7a31"
```

Three things just happened, in order:

1. The data frame was serialized to parquet and uploaded **to S3** -- not to
   GitHub.
2. `metadata.json` and `version_history.json` were updated in the git clone and
   committed.
3. The metadata commit was pushed to GitHub. The version is auditable from any
   machine with repo access, while the raw data stays in your bucket.

## Read it back as a reader

Writing is a developer action. Reading is what analysts and downstream
pipelines do, using the **reader role** -- a connection built from the bucket
credentials alone, with no GitHub PAT and no local clone. Leave `github_pat`
unset and connect by project name:

```{r}
reader_store <- datom_store(
  governance = NULL,
  data       = datom_store_s3(
    bucket     = bucket,
    prefix     = prefix,
    region     = region,
    access_key = access_key,
    secret_key = secret_key
  )
)                                         # no PAT -> reader role

reader_conn <- datom_get_conn(store = reader_store, project_name = project_name)

dm_back <- datom_read(reader_conn, "dm")
identical(dm_back, tibble::as_tibble(dm_m1))
#> [1] TRUE
```

datom stores data in [Apache Parquet](https://parquet.apache.org/) and reads it
back as a `tibble`. The read streams the parquet object from S3 directly; it
does **not** go through GitHub. A teammate on another machine takes the same
path -- they need access to the bucket, not to the git repo.

Inspect the version history at any time (developer connection):

```{r}
datom_status(conn)        # table list with SHAs
datom_history(conn, "dm") # version history
```

## Where you are

You have a fully versioned datom project running directly on S3:

- One table (`dm`) with one version, one parquet object in your bucket
- Metadata committed and pushed to GitHub -- version history is auditable
- Parquet data lives in S3 -- **nothing sensitive went to GitHub**
- No governance attached; the project stands on its own `project.yaml`

From here, the monthly-cadence workflow in [Month 2 Arrives](month-2-arrives.html)
and the bulk-import workflow in [A Folder of Extracts](folder-of-extracts.html)
apply unchanged -- they are written against a local store, but every function
call is identical on S3. To trace how one table descends from another, see
[Tracing Data Lineage](source-lineage.html).

## Governance and migration come later

Two capabilities are deliberately **not** part of this starting story:

- **Governance** -- a shared portfolio register, dispatch-based reader access,
  and managed teardown. It is opt-in and provided by the governance companion
  package (`datomanager`), adopted when a project graduates from solo to
  shared/managed. Starting on S3 does not commit you to it.
- **Managed migration** -- moving an existing project's data between backends
  (local to S3, or S3 to S3) while preserving history. That is a separate,
  governed workflow documented with the companion package. Because you started
  on S3, you do not need it now; it remains available later without
  contradicting anything you set up here.

In other words, **start-on-S3** and **migrate-to-S3** are complementary entry
points: this article is the greenfield path, and managed migration is the
move-an-existing-project path.

## Teardown

When you are done exploring, tear the project down. `datom_repo_delete()` removes
the GitHub metadata repo and the local clone. **Do this before** deleting the
local directory by hand -- removing the clone first strips the GitHub remote
reference and the remote repo will not be deleted.

```{r}
datom_repo_delete(conn, confirm = "STUDY_001")
```

`datom_repo_delete()` does **not** empty your S3 bucket -- bucket lifecycle is
your organization's domain, not datom's. Remove the project's objects from S3
with your own tooling (for example, the AWS CLI) if you no longer need them.
