---
title: "Getting Started with tbit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tbit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

tbit provides version-controlled data management for reproducible workflows. It
stores tabular data in S3 as parquet files while tracking metadata in git,
giving you:
 
- **Versioned data**: every write creates a traceable version
- **Reproducibility**: read any historical version by its SHA
- **Separation of concerns**: git tracks *what changed*; S3 stores *the data*
- **Two user roles**: *developers* create and update data; *readers* consume it

This vignette walks through both workflows end-to-end.

## Concepts

| Concept | Description |
|---------|-------------|
| **tbit repo** | A git repository with a `.tbit/` config directory |
| **Table** | A named dataset stored as parquet in S3 with versioned metadata in git |
| **`data_sha`** | SHA-256 of the parquet data (content-addressed storage) |
| **`metadata_sha`** | SHA-256 of the sorted metadata (used as version identifier) |
| **Manifest** | `.tbit/manifest.json` — catalog of all tables and their current state |
| **`input_files/`** | Flat directory for source data files (CSV, Excel, SAS, etc.) — gitignored |

## Setup

### Credentials

tbit derives credential environment variable names from the project name.
For a project called `"clinical_data"`, set:

```{r}
# AWS credentials for S3 access
Sys.setenv(
  TBIT_CLINICAL_DATA_ACCESS_KEY_ID = "your_access_key",
  TBIT_CLINICAL_DATA_SECRET_ACCESS_KEY = "your_secret_key"
)

# GitHub PAT — required for developers (git push/pull)
Sys.setenv(GITHUB_PAT = "ghp_your_token_here")
```

The naming convention is `TBIT_{PROJECT_NAME}_ACCESS_KEY_ID` and 
`TBIT_{PROJECT_NAME}_SECRET_ACCESS_KEY`, where the project name is uppercased
with spaces converted to underscores.

## Data Developer Workflow

Data developers have full access: they create repositories, write data,
and manage the git + S3 lifecycle.

### 1. Initialize a repository

```{r}
library(tbit)

tbit_init_repo(
  path         = "my_project",
  project_name = "clinical_data",
  remote_url   = "https://github.com/org/data-repo.git",
  bucket       = "shared-data-bucket",
  prefix       = "project-alpha/",
  region       = "us-east-1"
)
```

This creates:

- `.tbit/project.yaml` — project configuration
- `.tbit/routing.json` — read method routing
- `.tbit/manifest.json` — empty table catalog
- `.gitignore` — excludes data files and `input_files/`
- `input_files/` — directory for source data

The repo is initialized with git, connected to the remote, and the initial
commit is pushed.

### 2. Connect

```{r}
conn <- tbit_get_conn(path = "my_project")
conn
#> <tbit_conn>
#>   project: clinical_data
#>   bucket:  shared-data-bucket
#>   prefix:  project-alpha/
#>   role:    developer
```

### 3. Add data files

Place source files in the `input_files/` directory. This directory must be 
**flat** — no subdirectories allowed. tbit supports any format that
[rio](https://cran.r-project.org/package=rio) can import: CSV, TSV, Excel
(`.xlsx`), SAS (`.sas7bdat`), SPSS (`.sav`), Stata (`.dta`), parquet, and more.

```
input_files/
├── customers.csv
├── orders.tsv
└── lab_results.sas7bdat
```

The table name is the filename without extension (e.g., `customers.csv` →
table `"customers"`).

### 4. Scan the manifest

Before syncing, scan to see what's new or changed:

```{r}
manifest <- tbit_sync_manifest(conn)
#> ℹ Scanned 3 files: 3 new, 0 changed, 0 unchanged.

manifest
#>            name                    file    format   file_sha status
#> 1     customers input_files/customers.csv    csv   a1b2c3... new
#> 2        orders input_files/orders.tsv       tsv   d4e5f6... new
#> 3   lab_results input_files/lab_results.sas7bdat sas7bdat 7g8h9i... new
```

Use the `pattern` argument to filter:

```{r}
# Only CSV files
manifest_csv <- tbit_sync_manifest(conn, pattern = "*.csv")
```

### 5. Sync to S3

Process the manifest — each file is imported, converted to parquet, uploaded
to S3, and committed to git:

```{r}
results <- tbit_sync(conn, manifest)
#> ℹ Syncing 3 tables...
#> ✔ "customers" synced (new).
#> ✔ "orders" synced (new).
#> ✔ "lab_results" synced (new).
#> ℹ Sync complete: 3 succeeded, 0 failed, 0 skipped.

results[, c("name", "status", "result")]
#>          name status  result
#> 1   customers    new success
#> 2      orders    new success
#> 3 lab_results    new success
```

If a file fails (e.g., corrupt format), the default `continue_on_error = TRUE`
keeps going. Failed tables show `result = "error"` with details in the `error`
column.

### 6. Write individual tables

For programmatic workflows, write a data frame directly:

```{r}
# Create or update a table
new_data <- data.frame(
  patient_id = 1:100,
  treatment  = sample(c("A", "B"), 100, replace = TRUE),
  outcome    = rnorm(100)
)

tbit_write(conn, data = new_data, name = "trial_results",
           message = "Initial trial data")
#> ✔ Wrote "trial_results" (full): "f3a8b2c1"
```

tbit detects changes automatically. If you write the same data again, it's a
no-op:

```{r}
tbit_write(conn, data = new_data, name = "trial_results")
#> ℹ No changes detected for "trial_results". Skipping write.
```

### 7. Check status

```{r}
tbit_status(conn)
#> ── tbit status ──
#> ℹ Project: "clinical_data"
#> ℹ Bucket: "shared-data-bucket"
#> ℹ Role: "developer"
#> ℹ Tables on S3: 4
#> ✔ Git: clean (no uncommitted changes)
#> ℹ Branch: "main"
#> ✔ Input files: all 3 unchanged
```

### 8. Validate consistency

Verify that git metadata and S3 storage are in sync:

```{r}
result <- tbit_validate(conn)
#> ✔ All checks passed. Git and S3 are consistent.

result$valid
#> [1] TRUE
```

If there are mismatches (e.g., after a migration), auto-fix them:

```{r}
result <- tbit_validate(conn, fix = TRUE)
#> ⚠ Found 2 issues: 1 repo-level, 1 table-level.
#> ℹ Attempting to fix by syncing metadata to S3...
#> ✔ Fix applied. Re-run `tbit_validate()` to verify.
```

## Data Reader Workflow

Data readers only need S3 credentials — no git, no local repo, no `git2r`.

### 1. Connect

```{r}
library(tbit)

# Credentials (provided by the data developer)
Sys.setenv(
  TBIT_CLINICAL_DATA_ACCESS_KEY_ID = "reader_access_key",
  TBIT_CLINICAL_DATA_SECRET_ACCESS_KEY = "reader_secret_key"
)

conn <- tbit_get_conn(
  bucket       = "shared-data-bucket",
  prefix       = "project-alpha/",
  project_name = "clinical_data"
)
```

### 2. List available tables

```{r}
tables <- tbit_list(conn)
tables
#>          name current_version  last_updated version_count
#> 1   customers       a1b2c3...  2026-02-15             3
#> 2      orders       d4e5f6...  2026-02-15             1
#> 3 lab_results       7g8h9i...  2026-02-14             5
```

Filter with glob patterns:

```{r}
tbit_list(conn, pattern = "lab_*")
```

### 3. Read data

```{r}
# Latest version
customers <- tbit_read(conn, "customers")

# Specific version (metadata_sha from version history)
customers_v1 <- tbit_read(conn, "customers", version = "a1b2c3d4e5f6...")
```

### 4. View version history

```{r}
history <- tbit_history(conn, "customers", n = 5)
history
#>    version   data_sha        author          message                  timestamp
#> 1 f3a8b2.. abc123.. Jane Doe  Updated demographics 2026-02-15T10:30:00Z
#> 2 e2d1c0.. abc123.. Jane Doe  Fix column names     2026-02-14T15:20:00Z
#> 3 d1c0b9.. 789def.. Jane Doe  Initial load         2026-02-13T09:00:00Z
```

Note that versions with the same `data_sha` mean the data didn't change — only
metadata was updated.

### 5. Check status

```{r}
tbit_status(conn)
#> ── tbit status ──
#> ℹ Project: "clinical_data"
#> ℹ Bucket: "shared-data-bucket"
#> ℹ Role: "reader"
#> ℹ Tables on S3: 4
```

## Advanced: Migration

When moving data to a new S3 bucket:

```{r}
# 1. Copy data externally
#    aws s3 sync s3://old-bucket/proj/ s3://new-bucket/proj/

# 2. Update project.yaml with new bucket info

# 3. Sync all routing metadata to the new location
conn <- tbit_get_conn(path = "my_project")
tbit_sync_routing(conn)
#> ⚠ This will update routing metadata for 147 tables.
#> ℹ Current location: s3://new-bucket/proj/tbit/
#> Proceed? [y/N] y
#> ✔ Synced 3 repo-level files to S3.
#> ℹ Sync routing complete: 147 tables synced, 0 errors.
```

## Function Reference

### Connection & Init

| Function | Role | Description |
|----------|------|-------------|
| `tbit_init_repo()` | Developer | Initialize a new tbit repository |
| `tbit_get_conn()` | Both | Connect to a tbit repository |

### Read & Write

| Function | Role | Description |
|----------|------|-------------|
| `tbit_read()` | Both | Read a table (current or specific version) |
| `tbit_write()` | Developer | Write a data frame with change detection |

### Batch Sync

| Function | Role | Description |
|----------|------|-------------|
| `tbit_sync_manifest()` | Developer | Scan input files, detect new/changed |
| `tbit_sync()` | Developer | Process manifest, import and write all |
| `tbit_sync_routing()` | Developer | Push all git metadata to S3 |

### Query

| Function | Role | Description |
|----------|------|-------------|
| `tbit_list()` | Both | List tables from S3 manifest |
| `tbit_history()` | Both | Show version history for a table |
| `tbit_status()` | Both | Show connection, git, and sync state |

### Validation

| Function | Role | Description |
|----------|------|-------------|
| `tbit_validate()` | Developer | Check git-S3 consistency, optionally fix |
| `is_valid_tbit_repo()` | Both | Check if a path is a valid tbit repo |
