---
title: "Team Collaboration with datom"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Team Collaboration with datom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

datom supports multiple developers working on the same data repository concurrently.
Git tracks metadata; S3 stores data. This separation means developers can work in
parallel as long as they follow a simple discipline: **pull before you work, push
when you're done.**

> **Prerequisites**: This vignette assumes credentials are already configured.
> See `vignette("credentials")` for setup instructions including AWS keys,
> `github_pat`, and store construction.

This vignette covers:

- One-time repository setup (who runs `datom_init_repo()`)
- How teammates join an existing repository
- The daily pull → receive → sync workflow
- What to do when sync is rejected (push conflicts)
- Setting up read-only access for analysts

## Roles

datom recognizes two user types:

| Role | Git access | S3 access | Typical user |
|------|-----------|-----------|--------------|
| **Developer** | Read + write | Read + write | Data engineer, receives extracts, runs `datom_sync()` |
| **Reader** | None | Read only | Statistician, analyst, reads versioned tables |

Developers have a local clone of the git repo. Readers connect directly to S3
using `datom_get_conn()` with a store and project name.

## One-Time Setup: Initializing the Repository

**One person** initializes the repository. Everyone else clones it.

```{r}
library(datom)

# Build a store (see vignette("credentials") for details)
s3 <- datom_store_s3(
  bucket     = "clinical-data-bucket",
  prefix     = "study-001/",
  access_key = keyring::key_get("AWS_ACCESS_KEY"),
  secret_key = keyring::key_get("AWS_SECRET_KEY")
)

store <- datom_store(
  governance = s3,
  data       = s3,
  github_pat = keyring::key_get("GITHUB_PAT")
)

# The lead data engineer runs this once
datom_init_repo(
  path         = "study_001_data",
  project_name = "STUDY_001",
  store        = store,
  create_repo  = TRUE,
  repo_name    = "study-001-data"
)
```
```
#> ✔ Created GitHub repo "study-001-data"
#> ✔ Initialized datom repository at study_001_data
#> ✔ Pushed to origin/main
#> ✔ Synced routing.json and manifest.json to S3
```

This creates the folder structure, creates a GitHub repo, pushes the initial
commit, and uploads `routing.json` and `manifest.json` to S3.

### S3 Namespace Safety

`datom_init_repo()` checks whether the S3 namespace is already occupied before
writing any files. If another project has already claimed
`s3://clinical-data-bucket/study-001/datom/.metadata/manifest.json`, you'll get
a clear error:

```{r}
datom_init_repo(
  path         = "my_project",
  project_name = "MY_PROJECT",
  store        = store
)
```
```
#> ✖ S3 namespace already occupied by project "STUDY_001".
#>   Location: s3://clinical-data-bucket/study-001/datom/
#> ℹ Use a different bucket or prefix, or pass `.force = TRUE` to override.
```

This prevents accidental overwrites when multiple teams share an S3 bucket.

## Joining an Existing Repository

Every other team member clones with `datom_clone()`. They need a store with
the `remote_url` of the existing repo:

```{r}
# Teammate builds their own store with the remote URL
store <- datom_store(
  governance = datom_store_s3(
    bucket     = "clinical-data-bucket",
    prefix     = "study-001/",
    access_key = keyring::key_get("AWS_ACCESS_KEY"),
    secret_key = keyring::key_get("AWS_SECRET_KEY")
  ),
  data = datom_store_s3(
    bucket     = "clinical-data-bucket",
    prefix     = "study-001/",
    access_key = keyring::key_get("AWS_ACCESS_KEY"),
    secret_key = keyring::key_get("AWS_SECRET_KEY")
  ),
  github_pat = keyring::key_get("GITHUB_PAT"),
  remote_url = "https://github.com/org/study-001-data.git"
)

conn <- datom_clone(path = "study_001_data", store = store)
```
```
#> ✔ Cloned "STUDY_001" to study_001_data
```

```{r}
print(conn)
```
```
#> ── datom connection
#> • Project: "STUDY_001"
#> • Role: "developer"
#> • Bucket: "clinical-data-bucket"
#> • Prefix: "study-001/"
#> • Region: "us-east-1"
#> • Path: /home/bob/study_001_data
```

`datom_clone()` wraps `git2r::clone()` and returns a ready-to-use connection —
no re-initialization needed.

## Daily Workflow

### Step 1: Pull Latest Changes

Every work session starts with `datom_pull()`:

```{r}
conn <- datom_get_conn(path = "study_001_data")

datom_pull(conn)
```
```
#> ✔ Pulled 3 commits on "main".
```

`datom_pull()` fetches and merges upstream git commits — including any metadata
changes from teammates (manifest updates, new table versions, routing changes).
Git is the source of truth for all metadata.

If you're already up to date:
```
#> ℹ Already up to date on "main".
```

> **Note**: You don't strictly need to call `datom_pull()` manually before every
> `datom_sync()`. datom's push mechanism automatically pulls before pushing to
> ensure you're up to date. However, calling `datom_pull()` at the start of a
> session is still recommended — it lets you see what teammates have changed
> before you begin your own work.

### Step 2: Receive Data and Sync

Place new data extracts in `input_files/`, then **scan** and **sync**. These are
intentionally separate steps:

```{r}
# Example: write a data extract to the input directory
dm <- datom_example_data("dm", cutoff_date = "2026-03-28")
write.csv(dm, "study_001_data/input_files/dm.csv", row.names = FALSE)

# Scan — detects new and changed files in input_files/
manifest <- datom_sync_manifest(conn)

# Inspect before syncing — what did datom find?
manifest
```
```
#>   table_name status      file_path
#> 1         dm changed  input_files/dm.csv
```

The scan step produces a manifest that you can inspect before committing to a
sync. This is your checkpoint: verify that the right files were detected and
that statuses (new, changed, unchanged) make sense before writing anything to
git or S3.

```{r}
# Sync — imports, converts to parquet, writes to git + S3
results <- datom_sync(conn, manifest)
```
```
#> ℹ Syncing 1 table...
#> ✔ dm synced (changed).
#> ℹ Sync complete: 1 succeeded, 0 failed, 0 skipped.
```

> **Why two steps?** Separating scan from sync gives you a chance to review
> what will be synced before anything is written. In a team setting this is
> especially valuable — you can catch stale files, accidental duplicates, or
> unexpected changes before they hit git and S3. Future versions of datom will
> expand manifest-level APIs for richer pre-sync inspection (e.g., adding
> table descriptions, tagging QA status).

### Step 3: Verify

```{r}
# Check repository consistency
datom_validate(conn)
```
```
#> ✔ All checks passed. Git and S3 are consistent.
```

`datom_validate()` confirms that your local git state and S3 agree — metadata
matches data, all referenced parquet files exist in S3, and nothing is missing
or orphaned. Run it after every sync session.

That's it. The sync committed to git, pushed, and uploaded data + metadata to S3.
Your teammates can now `datom_pull()` to see the new version.

## Handling Conflicts

### Scenario: Push Rejected

If a teammate pushed while you were syncing, `datom_sync()` will abort with a
clear message:

```
#> ✖ Local git branch "main" is behind remote by 2 commits.
#> ℹ Run `datom_pull()` or `git pull` to update before syncing.
```

**Resolution:**

```{r}
# 1. Pull the latest changes
datom_pull(conn)

# 2. Re-scan (manifest may have changed)
manifest <- datom_sync_manifest(conn)

# 3. Re-sync
results <- datom_sync(conn, manifest)
```

### Scenario: Merge Conflict

Merge conflicts occur when two developers modify the same file (e.g., both
update `metadata.json` for the same table). This is rare in practice because
datom auto-generates metadata files — but it can happen if developers sync the
same table simultaneously.

```
#> ✖ Merge conflict detected — manual resolution required.
#> ℹ Pull latest changes, resolve conflicts, and re-run.
#> ℹ Use `git status` to see conflicting files.
```

**Resolution:**

1. Open a terminal in the repo directory
2. Run `git status` to see conflicting files
3. Edit the files to resolve conflicts (or accept one side with `git checkout --theirs`)
4. `git add` the resolved files and `git commit`
5. Re-run `datom_sync()`

> **Tip**: Merge conflicts are minimized when each developer is responsible for
> different tables. Coordinate who syncs which tables within the team.

## Setting Up Readers

Readers (analysts, statisticians) don't need git access. They connect directly
to S3 using a reader store:

```{r}
# Reader store — no github_pat
reader_s3 <- datom_store_s3(
  bucket     = "clinical-data-bucket",
  prefix     = "study-001/",
  access_key = keyring::key_get("AWS_ACCESS_KEY_READER"),
  secret_key = keyring::key_get("AWS_SECRET_KEY_READER")
)

reader_store <- datom_store(governance = reader_s3, data = reader_s3)

reader_conn <- datom_get_conn(store = reader_store, project_name = "STUDY_001")
print(reader_conn)
```
```
#> ── datom connection
#> • Project: "STUDY_001"
#> • Role: "reader"
#> • Bucket: "clinical-data-bucket"
#> • Prefix: "study-001/"
#> • Region: "us-east-1"
```

Readers can list tables, read data, and view history — all via S3:

```{r}
datom_list(reader_conn)
datom_read(reader_conn, "dm")
datom_history(reader_conn, "dm")
```

### What to Share with Readers

Provide your readers with:

| Item | Example |
|------|---------|
| `project_name` | `"STUDY_001"` |
| `bucket` | `"clinical-data-bucket"` |
| `prefix` | `"study-001/"` |
| AWS credentials | Reader-scoped IAM keys (read-only S3 access) |

For credential setup details, see `vignette("credentials")`.

## Recommended Team Practices

### 1. One person initializes, everyone else clones

```{r}
# Alice (lead):
datom_init_repo(
  path = "study_data", project_name = "STUDY_001",
  store = store, create_repo = TRUE, repo_name = "study-001-data"
)

# Bob and Carol (build store with remote_url, then clone):
conn <- datom_clone(path = "study_data", store = store)
```

### 2. Pull at the start of every session

```{r}
# First line of every script
conn <- datom_get_conn(path = "study_data")
datom_pull(conn)
```

### 3. Divide tables across developers

If Alice syncs `dm` and `lb`, and Bob syncs `ex` and `vs`, merge conflicts are
virtually impossible.

### 4. Validate after sync

```{r}
datom_validate(conn)
```

This confirms git and S3 are consistent. Run it after every sync session, and
especially after resolving conflicts.

### 5. Use `datom_status()` to see what's changed

```{r}
datom_status(conn)
```

Before syncing, check which tables have pending changes. This helps avoid
syncing stale data or re-syncing unchanged tables.

## Summary

| Step | Command | Who |
|------|---------|-----|
| Build store | `datom_store(...)` | Everyone (once per session) |
| Initialize (once) | `datom_init_repo(store = ...)` | Lead developer |
| Clone | `datom_clone(store = ...)` | All other developers |
| Start session | `datom_pull(conn)` | All developers |
| Receive data | Place files in `input_files/` | Developer |
| Scan | `datom_sync_manifest(conn)` | Developer |
| Sync | `datom_sync(conn, manifest)` | Developer |
| Validate | `datom_validate(conn)` | Developer |
| Read data | `datom_read(conn, "table")` | Developer or reader |
| Conflict? | `datom_pull()` → re-scan → re-sync | Developer |
