---
title: "Buckets and Prefixes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Buckets and Prefixes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

How you map studies and data products onto S3 buckets and prefixes shapes
your IAM story, your retention policy, and (eventually) your migration
costs. This article describes the **convention datom recommends** for new
projects, explains why, and shows when to deviate.

> **datom enforces nothing here.** `bucket` and `prefix` are plain strings
> that produce keys of the form `{prefix}/datom/{project_name}/...`. The
> patterns below are *conventions* -- ways to organize the namespace that
> make access control and lifecycle management cleaner. Pick what fits
> your organization; the package supports anything.

## The recommended convention: bucket-per-study

For most clinical-trial work, use **one bucket per study**, with **prefixes
distinguishing data products**:

```
study-001-datom/                          # one bucket = one study
    datom/STUDY_001/...                   # raw data (prefix = "")
    adam/datom/ADAM_STUDY_001/...         # ADaM (prefix = "adam/")
    tlf/datom/TLF_STUDY_001/...           # TLFs (prefix = "tlf/")
```

In code:

```{r}
# Raw clinical extracts
raw_store <- datom_store_s3(
  bucket     = "study-001-datom",
  prefix     = "",                        # raw lives at the bucket root
  region     = "us-east-1",
  access_key = ..., secret_key = ...
)

# ADaM data product (derived)
adam_store <- datom_store_s3(
  bucket     = "study-001-datom",
  prefix     = "adam/",
  region     = "us-east-1",
  access_key = ..., secret_key = ...
)
```

### Why this works

1. **Bucket = study = IRB unit.** A study is the natural unit of consent,
   audit, and access governance. Aligning the storage boundary with the
   governance boundary makes "grant a reader access to STUDY-001" a
   one-sentence IAM policy.

2. **Prefix = data product.** Each derived product (ADaM, TLF, custom
   pools) lives under its own prefix. Lifecycle rules apply naturally:
   raw stays for the full retention window, intermediate products can
   sunset earlier.

3. **It scales gracefully.** When you outgrow it (many studies under
   one program, IT pressure to consolidate buckets), the upgrade path
   to **program-per-bucket** is mechanical: replace the empty prefix
   with `study-XYZ/`. No datom code change required.

4. **It plays well with the decommission boundary.** `datom_decommission()`
   removes the `datom/{project}/` namespace inside the bucket -- the
   bucket itself is caller-owned. With Pattern A the user mental model
   is "the bucket is the study", and the boundary is intuitive.

### Sub-study visibility (subset of prefixes)

A common requirement: an analyst should see ADaM but not raw clinical
data. Three mechanisms work with Pattern A, none of which require any
datom code change:

1. **IAM policy with `Resource` ARN + prefix.** Attach to the analyst's
   role:

    ```json
    {
      "Effect": "Allow",
      "Action": ["s3:GetObject", "s3:ListBucket"],
      "Resource": [
        "arn:aws:s3:::study-001-datom",
        "arn:aws:s3:::study-001-datom/adam/*"
      ],
      "Condition": {
        "StringLike": {"s3:prefix": ["adam/*"]}
      }
    }
    ```

    The analyst constructs `datom_store_s3(bucket="study-001-datom",
    prefix="adam/")`; attempts to read raw return 403.

2. **S3 Access Points** -- create an access point `adam-readers` on
   `study-001-datom` with a prefix policy restricting reads to `adam/`.
   datom carries an optional `endpoint` field on `datom_conn` for
   exactly this case; the planned governance companion package will
   provision access points automatically.

3. **Lake Formation / bucket policies** for organizations already
   running fine-grained data governance.

The same mechanism applies to the **gov bucket** for per-project
visibility: a reader's gov credentials can be scoped to
`projects/STUDY_001/*`.

## Governance bucket convention

Governance is **organization-scoped by design**: one gov repo, many
projects registered under `projects/{name}/`. Recommend:

```
acme-datom-gov/                           # one bucket per organization
    projects/STUDY_001/...
    projects/ADAM_STUDY_001/...
    projects/STUDY_002/...
```

In code:

```{r}
gov_store <- datom_store_s3(
  bucket     = "acme-datom-gov",
  prefix     = "",                        # dedicated bucket; no prefix needed
  region     = "us-east-1",
  access_key = ..., secret_key = ...
)
```

Use a **dedicated bucket with empty prefix** when you can. Use a prefix
(e.g. `prefix = "datom-gov/"`) only when you must share the bucket with
other tooling -- the prefix then namespaces datom inside a multi-tenant
bucket.

## Alternative patterns

### Program-per-bucket

When a single compound has many studies (multiple phases, indications,
populations), consolidate at the program level:

```
compound-X-datom/
    study-001/datom/STUDY_001/...         # prefix = "study-001/"
    study-001/adam/datom/ADAM_STUDY_001/...
    study-002/datom/STUDY_002/...
    study-002/adam/datom/ADAM_STUDY_002/...
```

- **IAM**: scope at prefix level (`study-001/*`) rather than bucket level.
  More policy work; fewer buckets.
- **When to use**: 20+ studies under one program; IT pressure on bucket
  counts; consistent IRB scope per program.
- **Migration from Pattern A**: each study's bucket becomes a top-level
  prefix in the program bucket. Move data once via the future
  `datom_migrate_data()`; `ref.json` is rewritten and readers pick up
  the new location automatically.

### Org-per-bucket

One bucket for the whole organization, everything prefixed:

```
acme-datom/
    program-X/study-001/datom/STUDY_001/...
    program-Y/study-042/datom/STUDY_042/...
```

- **IAM**: requires policy-document discipline. Only viable with mature
  IAM tooling (Lake Formation, SCPs, OPA).
- **When to use**: large organizations with centralized data governance,
  where the bucket-count overhead of Pattern A becomes a real cost.
- **Trade-off**: highest flexibility, weakest natural boundaries.

## Migration between patterns

The patterns are not mutually exclusive in time. A common path:

1. Start with **Pattern A** (bucket-per-study). Simple, fast, IRB-aligned.
2. Around 20-30 studies, IT requests bucket consolidation. Migrate to
   **Pattern B** (program-per-bucket).
3. Eventually, large orgs centralize on **Pattern C** with Lake Formation.

datom's `ref.json` is the migration commit point: once gov updates the
location, readers automatically follow. Per-version history is preserved
when `datom_migrate_data()` ships; until then, migrations decommission
+ re-init (see [S3 Setup and Promotion](promoting-to-s3.html) for the
mechanics).

## Quick decision guide

| You have... | Use |
|---|---|
| One study, small team, no special IAM needs | Pattern A, empty data prefix |
| Multiple data products per study (raw + ADaM + TLF) | Pattern A, prefix per product |
| Many studies under one compound, IT wants fewer buckets | Pattern B |
| Enterprise data lake with central governance | Pattern C |
| Sharing the gov bucket with other tooling | Pattern A or B with `prefix = "datom-gov/"` |

When in doubt: **start with Pattern A**. You can migrate later; the
inverse is rarely true.
